---
output: github_document
---

<!-- README.md is generated from README.Rmd using [duckknit](https://github.com/rundel/duckknit). Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(duckknit)
knitr::opts_chunk$set(mode = "table", comment = "")

# Rewrite duckknit's ``` duckdb fences to ``` sql so GitHub applies
# SQL syntax highlighting in the rendered README.
knitr::knit_hooks$set(document = function(x) {
  gsub("(?m)^``` duckdb", "``` sql", x, perl = TRUE)
})

# Pre-start the default duckknit session with -unsigned so unsigned extensions
# can be loaded. duckknit_start_session() only accepts a db path, not extra
# CLI flags, so we launch the process directly and register it in duckknit's
# internal session table.
options(duckknit.duckdb = "/root/.local/bin/duckdb")
local({
  ns  <- asNamespace("duckknit")
  cmd <- duckknit:::find_duckdb()
  proc <- processx::process$new(
    command = cmd,
    args = c("-unsigned", ":memory:"),
    stdin = "|", stdout = "|", stderr = "|",
    cleanup = TRUE, cleanup_tree = TRUE
  )
  unlockBinding(".sessions", ns)
  ns$.sessions[["readme"]] <- proc
  lockBinding(".sessions", ns)
  unlockBinding(".state", ns)
  ns$.state$last_session   <- "readme"
  lockBinding(".state", ns)
})
```

```{duckdb, include = FALSE}
LOAD 'build/release/ducknng.duckdb_extension';
```

# ducknng

`ducknng` is a pure C DuckDB extension that exposes DuckDB SQL and manifest-declared RPC over [NNG](https://nng.nanomsg.org/) transports and HTTP/HTTPS carriers. A SQL session can call another DuckDB session across `inproc://`, `ipc://`, `tcp://`, `tls+tcp://`, `ws://`, `wss://`, `http://`, or `https://`, and RPC row payloads ride on Arrow IPC.

It draws on two R packages. [`nanonext`](https://github.com/r-lib/nanonext) supplies the socket and AIO ergonomics; [`mangoro`](https://github.com/sounkou-bioinfo/mangoro) supplies the thin versioned envelope. The DuckDB integration keeps generated SQL injection-safe but assumes the deployment owns the trust boundary — see **Deployment and admission policy** below.

### How the pieces stack

The extension is one runtime with four layers; later sections walk through each in turn.

1. **Transport.** NNG sockets and listeners across the full transport family. Synchronous send/recv plus first-class one-shot AIO handles. Per-service pipe-event telemetry (`ducknng_read_monitor(...)`, `ducknng_list_pipes(...)`).
2. **Framed RPC.** A versioned envelope carries Arrow IPC table payloads or JSON control text. Servers come up with `ducknng_start_server(...)`, and the listen URL scheme chooses NNG or HTTP/HTTPS. HTTP/HTTPS stays on the same server entrypoint with `contexts = 1`. Clients call by URL; synchronous helpers route by scheme, AIO helpers expose the same calls as futures.
3. **Policy.** Manifest-driven method registry with opt-in `exec`. Fast C admission for required mTLS, exact peer-identity allowlists, IP/CIDR allowlists, and service limits. Optional SQL authorizers run at the request boundary and read context through `ducknng_auth_context()`. Sessions are bearer-token owned, additionally bound to verified mTLS identity when present. The default execution model is `shared_serialized_connection`, with `service_serialized_connection` and `request_connection` available through `ducknng_set_service_execution_model(...)`; the current model is exposed in `manifest` and `ducknng_list_servers()`.
4. **Codecs.** `ducknng_parse_body(...)` and `ducknng_ncurl_table(...)` parse content-type-tagged BLOBs. Built-in providers cover JSON, Arrow IPC, ducknng frames, and a text/raw fallback. Deployments can override or extend the set with `ducknng_register_codec(content_type, function_name)`; the function takes `BLOB` and returns `VARCHAR`.

`ducknng` is intentionally low-level. Long-lived runtime handles — servers, sockets, AIO, TLS configs, query sessions — are **manually managed**. Stop, close, or drop them explicitly; runtime teardown is fallback cleanup, not the primary lifecycle.

## Quick tour

Build the extension:

```sh
make configure
make release
```

Load it, start an IPC service, ask the server its name through the built-in `manifest` method, and stop it again:

```{duckdb collapse=TRUE, comment=""}
-- args: name, listen URL, REP-context count, recv_max_bytes,
--       session_idle_ms, tls_config_id (0 = plaintext).
SELECT ducknng_start_server(
  'hello', 'ipc:///tmp/ducknng_hello.ipc',
  1, 134217728, 300000, 0::UBIGINT
);

-- The manifest column is JSON text held as VARCHAR; cast to JSON to drill in.
SELECT json_extract_string(manifest::JSON, '$.server.name') AS server_name,
       json_array_length(json_extract(manifest::JSON, '$.methods')) AS methods
FROM ducknng_get_rpc_manifest('ipc:///tmp/ducknng_hello.ipc', 0::UBIGINT)
WHERE ok;

SELECT ducknng_stop_server('hello');
```

Point `ducknng_start_server(...)` at an `http://` or `https://` URL with `contexts = 1` and the same RPC methods are mounted on the HTTP carrier. The **Examples** section walks through transport patterns, AIO, the client surface, body codecs, and sessions in that order.

## Lifetime and manual cleanup

DuckDB's extension API does give `ducknng` destroy callbacks for internal function-registration, bind, and init state. What DuckDB SQL does **not** give this project is a general-purpose GC/finalizer model for long-lived SQL handles like R external pointer finalizers. So at the public SQL surface, long-lived handles are manually managed.

You should explicitly clean up what you create:

- stop servers with `ducknng_stop_server(...)`
- close sockets with `ducknng_close_socket(...)`
- drop aio handles with `ducknng_aio_drop(...)`
- drop TLS config handles with `ducknng_drop_tls_config(...)`
- close query sessions with `ducknng_close_query(...)`

Important details:

- `ducknng_aio_cancel(...)` is cancellation control, not the destructor
- `ducknng_cancel_query(...)` is best-effort session control, not a replacement for the normal explicit close path
- runtime teardown is fallback cleanup, not the primary lifecycle model for a long-lived DuckDB process

For the deeper write-ups see `docs/lifetime.md`, `docs/protocol.md`, `docs/manifest.md`, `docs/security.md`, `docs/registry.md`, `docs/transports.md`, `docs/http.md`, `docs/http_server_framework.md`, `docs/codecs.md`, and `docs/types.md`. A multi-process subscriber gateway sketch now lives in `docs/subscriber_gateway_demo.md`; the older non-sealed routing note remains in `docs/mesh_routing_demo.md`. `NEWS.md` summarizes landed changes. TLS supports both file-backed and in-memory PEM material (`ducknng_tls_config_from_files(...)`, `ducknng_tls_config_from_pem(...)`, `ducknng_self_signed_tls_config(...)`); `auth_mode = 2` enables required mTLS peer verification for `tls+tcp://`, `wss://`, and `https://` services.

## Function catalog

The generated catalog is the public SQL surface. Any loaded `ducknng__*` helper names are internal macro implementation details required by DuckDB's stable C API path and are not supported user APIs.

```{r function-catalog, echo=FALSE, results='asis'}
catalog <- paste(readLines("function_catalog/functions.md"), collapse = "\n")
knitr::asis_output(paste0(
  "<details>\n",
  "<summary><strong>Expand the full generated function catalog</strong></summary>\n\n",
  catalog,
  "\n\n</details>\n"
))
```

## Examples

### Start an IPC listener and inspect the registry

```{duckdb collapse=TRUE, comment=""}
SELECT ducknng_start_server(
  'sql0',                         -- service name
  'ipc:///tmp/ducknng_sql0.ipc', -- listen URL
  1,                              -- REP contexts
  134217728,                      -- recv_max_bytes
  300000,                         -- session_idle_ms
  0                               -- tls_config_id (0 means plaintext)
);

SELECT name, listen, contexts, running, execution_model, sessions
FROM ducknng_list_servers();

SELECT ducknng_stop_server('sql0');
```

### Request multiple REP contexts on one REP socket

```{duckdb collapse=TRUE, comment=""}
SELECT ducknng_start_server(
  'sql_multi',                    -- service name
  'ipc:///tmp/ducknng_sql_multi.ipc', -- listen URL
  3,                              -- REP contexts
  134217728,                      -- recv_max_bytes
  300000,                         -- session_idle_ms
  0                               -- tls_config_id (0 means plaintext)
);

SELECT name, contexts, running
FROM ducknng_list_servers()
WHERE name = 'sql_multi';

SELECT ducknng_stop_server('sql_multi');
```

### DuckDB can also act as a client

#### Start the server and inspect the default registry

The default RPC surface only exposes `manifest`. `exec` is opt-in.

```{duckdb collapse=TRUE, comment=""}
SELECT ducknng_start_server(
  'sql_client_demo',
  'ipc:///tmp/ducknng_sql_client_demo.ipc',
  1, 134217728, 300000, 0
);

WITH manifest_row AS (
  SELECT manifest
  FROM ducknng_get_rpc_manifest('ipc:///tmp/ducknng_sql_client_demo.ipc', 0::UBIGINT)
  WHERE ok
)
SELECT json_extract_string(manifest::JSON, '$.server.name') AS server_name,
       json_array_length(json_extract(manifest::JSON, '$.methods')) AS method_count,
       position('"name":"exec"' IN manifest) > 0 AS has_exec
FROM manifest_row;

SELECT name, family, response_mode, requires_auth, disabled
FROM ducknng_list_methods()
ORDER BY name;
```

#### Register `exec` and re-inspect the manifest

```{duckdb collapse=TRUE, comment=""}
SELECT ducknng_register_exec_method(false) AS registered_exec;

SELECT name, family, response_mode, requires_auth, disabled
FROM ducknng_list_methods()
ORDER BY name;

WITH manifest_row AS (
  SELECT manifest
  FROM ducknng_get_rpc_manifest('ipc:///tmp/ducknng_sql_client_demo.ipc', 0::UBIGINT)
  WHERE ok
)
SELECT json_extract_string(manifest::JSON, '$.server.name') AS server_name,
       json_array_length(json_extract(manifest::JSON, '$.methods')) AS method_count,
       position('"name":"exec"' IN manifest) > 0 AS has_exec
FROM manifest_row;
```

#### Run RPC statements and fetch rows

```{duckdb collapse=TRUE, comment=""}
SELECT * FROM ducknng_run_rpc(
  'ipc:///tmp/ducknng_sql_client_demo.ipc',
  'CREATE TABLE IF NOT EXISTS client_side_demo(i INTEGER)',
  0::UBIGINT
);

SELECT * FROM ducknng_run_rpc(
  'ipc:///tmp/ducknng_sql_client_demo.ipc',
  'INSERT INTO client_side_demo VALUES (10), (11)',
  0::UBIGINT
);

SELECT *
FROM ducknng_query_rpc(
  'ipc:///tmp/ducknng_sql_client_demo.ipc',
  'SELECT i, i > 10 AS gt_10 FROM client_side_demo ORDER BY i',
  0::UBIGINT
);
```

#### Arrow IPC type roundtrip

The row path carries temporal, decimal, list, and struct values.

```{duckdb collapse=TRUE, comment=""}
SELECT d = DATE '2024-01-02' AS date_ok,
       ts = TIMESTAMP '2024-01-02 03:04:05.123456' AS ts_ok,
       dec = 123.45::DECIMAL(10,2) AS decimal_ok,
       xs[2] IS NULL AND xs[3] = 3 AS list_ok,
       st.a = 7 AND st.b = 'bee' AS struct_ok
FROM ducknng_query_rpc(
  'ipc:///tmp/ducknng_sql_client_demo.ipc',
  'SELECT DATE ''2024-01-02'' AS d,
          TIMESTAMP ''2024-01-02 03:04:05.123456'' AS ts,
          123.45::DECIMAL(10,2) AS dec,
          [1, NULL, 3]::INTEGER[] AS xs,
          {''a'': 7::INTEGER, ''b'': ''bee''} AS st',
  0::UBIGINT
);
```

#### Body codec helpers

```{duckdb collapse=TRUE, comment=""}
SELECT provider, output
FROM ducknng_list_codecs()
ORDER BY provider;

SELECT a, b
FROM ducknng_parse_body(
  '[{"a":1,"b":"x"},{"a":2,"b":"y"}]'::BLOB,
  'application/json; charset=utf-8'
)
ORDER BY a;
```

#### Primitive socket layer

Low-level socket operations return a struct-shaped result: `ok`, `error`, `nng_error`, `nng_error_message`, `socket_id`, `payload`, `url`.

```{duckdb collapse=TRUE, comment=""}
SELECT (ducknng_open_socket('req')).socket_id;

SELECT (ducknng_dial_socket(
  1, 'ipc:///tmp/ducknng_sql_client_demo.ipc', 1000, 0::UBIGINT
)).ok;

SELECT * FROM ducknng_list_sockets();

-- Send the built-in manifest request frame through the open socket.
SELECT ok, error, nng_error_message, octet_length(payload) > 0 AS has_payload
FROM ducknng_request_socket(
  1::UBIGINT,
  from_hex('01000000000000000000000000000000000000000000'),
  1000
);

-- One-shot URL form: no persistent socket handle needed.
SELECT ok, error, nng_error_message, octet_length(payload) > 0 AS has_payload
FROM ducknng_request(
  'ipc:///tmp/ducknng_sql_client_demo.ipc',
  from_hex('01000000000000000000000000000000000000000000'),
  1000,
  0::UBIGINT
);
```

#### Raw frame helpers

The raw scalar forms return reply bytes as BLOBs; `ducknng_decode_frame(...)` projects the envelope fields and text payload.

```{duckdb collapse=TRUE, comment=""}
-- Raw bytes from a socket-scoped request (first 14 octets as hex).
SELECT substr(
  hex(ducknng_request_socket_raw(
    1,
    from_hex('01000000000000000000000000000000000000000000'),
    1000
  )),
  1, 28
);

-- Decode the manifest reply directly.
SELECT ok, version, type_name, name,
       position('"name":"exec"' IN payload_text) > 0 AS has_exec
FROM ducknng_decode_frame(
  ducknng_get_rpc_manifest_raw('ipc:///tmp/ducknng_sql_client_demo.ipc', 0::UBIGINT)
);

-- Decode an exec reply.
SELECT ok, type_name, name
FROM ducknng_decode_frame(
  ducknng_run_rpc_raw(
    'ipc:///tmp/ducknng_sql_client_demo.ipc',
    'CREATE TABLE IF NOT EXISTS client_side_demo(i INTEGER)',
    0::UBIGINT
  )
);

-- The generic raw request helper decodes the same way.
SELECT ok, version, type_name, name,
       position('"name":"exec"' IN payload_text) > 0 AS has_exec
FROM ducknng_decode_frame(
  ducknng_request_raw(
    'ipc:///tmp/ducknng_sql_client_demo.ipc',
    from_hex('01000000000000000000000000000000000000000000'),
    1000, 0::UBIGINT
  )
);
```

#### Cleanup

```{duckdb collapse=TRUE, comment=""}
SELECT (ducknng_close_socket(1)).ok;
SELECT ducknng_stop_server('sql_client_demo');
```

### Use `ducknng_ncurl()` against a local `nanonext` HTTPS server

A tiny local R `nanonext` HTTPS server is started during README rendering so this example runs without depending on the public internet. The core server definition is shown below because the transport helper and its carrier behavior are part of the example rather than hidden setup. The hidden setup only launches the script in the background and manages the temporary PID, log, and CA files needed while rendering. `ducknng_ncurl(...)` remains the low-level HTTP/HTTPS transport primitive even though the higher-level synchronous request, RPC, and session helpers now auto-route over `http://` and `https://` when the URL scheme selects that carrier.

```{r http-client-demo-setup, echo=FALSE, results='asis'}
library(nanonext)
pidfile <- "/tmp/ducknng_readme_http_demo.pid"
script <- "/tmp/ducknng_readme_http_demo.R"
logfile <- "/tmp/ducknng_readme_http_demo.log"
cafile <- "/tmp/ducknng_readme_http_demo_ca.pem"
if (file.exists(pidfile)) {
  pid <- suppressWarnings(as.integer(readLines(pidfile, warn = FALSE)))
  if (!is.na(pid)) tools::pskill(pid)
  unlink(pidfile)
}
http_demo_code <- paste(
  "library(nanonext)",
  "",
  "cert <- write_cert(cn = '127.0.0.1')",
  sprintf("writeLines(cert$client[[1]], %s)", deparse(cafile)),
  "",
  "server <- http_server(",
  "  url = 'https://127.0.0.1:18443',",
  "  handlers = list(",
  "    handler('/hello', function(req) {",
  "      list(",
  "        status = 200L,",
  "        headers = c(",
  "          'Content-Type' = 'text/plain',",
  "          'X-Test' = 'hello'",
  "        ),",
  "        body = 'hello from nanonext https server'",
  "      )",
  "    }),",
  "    handler('/echo', function(req) {",
  "      list(",
  "        status = 200L,",
  "        headers = c(",
  "          'Content-Type' = (",
  "            req$headers[['Content-Type']] %||% 'application/octet-stream'",
  "          ),",
  "          'X-Test' = 'echo'",
  "        ),",
  "        body = req$body",
  "      )",
  "    }, method = 'POST')",
  "  ),",
  "  tls = tls_config(server = cert$server)",
  ")",
  sep = "\n"
)
cat("```r\n", http_demo_code, "\n```\n\n", sep = "")
launch_code <- paste(
  http_demo_code,
  "",
  "library(later)",
  "attr(server, 'start')()",
  "repeat later::run_now(Inf)",
  sep = "\n"
)
writeLines(launch_code, script)
status <- system(sprintf(
  "Rscript %s > %s 2>&1 & echo $! > %s",
  shQuote(script),
  shQuote(logfile),
  shQuote(pidfile)
))
if (status != 0 || !file.exists(pidfile)) {
  stop("failed to start the README HTTP demo server")
}
pid <- suppressWarnings(as.integer(readLines(pidfile, warn = FALSE)))
if (is.na(pid) || system(sprintf("kill -0 %d >/dev/null 2>&1", pid)) != 0) {
  log_text <- if (file.exists(logfile)) paste(readLines(logfile, warn = FALSE), collapse = "\n") else ""
  stop("README HTTP demo server did not stay running\n", log_text)
}
ready <- FALSE
for (attempt in seq_len(20)) {
  client_tls <- tryCatch(
    if (file.exists(cafile)) tls_config(client = cafile) else NULL,
    error = function(e) NULL
  )
  probe <- tryCatch(
    if (!is.null(client_tls)) ncurl("https://127.0.0.1:18443/hello", timeout = 500L, tls = client_tls) else NULL,
    error = function(e) NULL
  )
  if (!is.null(probe) && identical(probe$status, 200L)) {
    ready <- TRUE
    break
  }
  Sys.sleep(0.25)
}
if (!ready) {
  log_text <- if (file.exists(logfile)) paste(readLines(logfile, warn = FALSE), collapse = "\n") else ""
  stop("README HTTP demo server did not become ready\n", log_text)
}
```

```{duckdb collapse=TRUE, comment=""}
-- Register a client TLS handle that trusts the self-signed HTTPS demo server.
SET VARIABLE tls_http_id = ducknng_tls_config_from_files(
  NULL,                                     -- cert_key_file
  '/tmp/ducknng_readme_http_demo_ca.pem',   -- ca_file
  NULL,                                     -- password
  2                                         -- auth_mode = require certificate validation
);
SELECT getvariable('tls_http_id') AS tls_config_id;

-- Call /hello through ducknng_ncurl(): NULL method means GET and NULL body means no request body.
SELECT ok, status, error, body_text
FROM ducknng_ncurl(
  'https://127.0.0.1:18443/hello', -- url
  NULL,                            -- method
  NULL,                            -- headers_json
  NULL,                            -- body
  2000,                            -- timeout_ms
  getvariable('tls_http_id')::UBIGINT  -- tls_config_id
);

-- POST can send raw bytes while still exposing HTTP headers and status in-band.
SET VARIABLE echo_headers =
  '[{"name":"Content-Type","value":"application/octet-stream"}]';

SELECT
  ok,
  status,
  error,
  hex(body) AS body_hex,
  position('X-Test' IN headers_json) > 0 AS has_header
FROM ducknng_ncurl(
  'https://127.0.0.1:18443/echo', -- url
  'POST',                         -- method
  getvariable('echo_headers'),    -- headers_json
  from_hex('01020304'),           -- body
  2000,                           -- timeout_ms
  getvariable('tls_http_id')::UBIGINT  -- tls_config_id
);

-- Drop the temporary client TLS handle after the HTTPS checks complete.
SELECT ducknng_drop_tls_config(getvariable('tls_http_id')::UBIGINT);
```

```{r http-client-demo-cleanup, include=FALSE}
pidfile <- "/tmp/ducknng_readme_http_demo.pid"
if (file.exists(pidfile)) {
  pid <- suppressWarnings(as.integer(readLines(pidfile, warn = FALSE)))
  if (!is.na(pid)) tools::pskill(pid)
  unlink(pidfile)
}
unlink(c(
  "/tmp/ducknng_readme_http_demo.R",
  "/tmp/ducknng_readme_http_demo.log",
  "/tmp/ducknng_readme_http_demo_ca.pem"
))
```

### Start `ducknng` on `http://` and use the routed helpers directly

`ducknng_start_server(...)` mounts the existing framed RPC surface at the exact path encoded in the listen URL, and the higher-level synchronous request, RPC, and session helpers keep the same names they already use for NNG URLs while switching carriers automatically from the scheme. For `http://` and `https://`, use `contexts = 1`.

```{duckdb collapse=TRUE, comment=""}
SELECT ducknng_start_server(
  'http_demo',
  'http://127.0.0.1:18444/_ducknng',
  1,
  134217728,
  300000,
  0::UBIGINT
);

WITH manifest_row AS (
  SELECT manifest
  FROM ducknng_get_rpc_manifest('http://127.0.0.1:18444/_ducknng', 0::UBIGINT)
  WHERE ok
)
SELECT json_extract_string(manifest::JSON, '$.server.name') AS server_name,
       json_array_length(json_extract(manifest::JSON, '$.methods')) AS method_count,
       position('"name":"manifest"' IN manifest) > 0 AS has_manifest
FROM manifest_row;

SET VARIABLE http_demo_frame = (
  SELECT body
  FROM ducknng_ncurl(
    'http://127.0.0.1:18444/_ducknng',
    'POST',
    '[{"name":"Content-Type","value":"application/vnd.ducknng.frame"}]',
    from_hex('01000000000000000000000000000000000000000000'),
    2000,
    0::UBIGINT
  )
);

SELECT ok, type_name, name
FROM ducknng_decode_frame(getvariable('http_demo_frame'));

SELECT ducknng_stop_server('http_demo');
```

### Register low-level HTTP routes beside the framed RPC mount

The framed RPC mount stays exactly where the listen URL says it lives, but HTTP and HTTPS services can now register additional exact, prefix, or template routes beside that mount. Route handlers are ordinary SQL queries that return exactly one response row, and they can inspect the in-flight request through `ducknng_http_request()` and `ducknng_http_request_body()`.

#### Start the server and register routes

```{duckdb collapse=TRUE, comment=""}
SELECT ducknng_start_server(
  'http_route_demo',
  'http://127.0.0.1:18445/_ducknng',
  1, 134217728, 300000, 0::UBIGINT
);

SELECT ducknng_register_http_route(
  'http_route_demo', 'GET', '/healthz',
  'SELECT * FROM ducknng_http_text(200, ''ok'')'
);

SELECT ducknng_register_http_route(
  'http_route_demo', 'POST', '/echo',
  'SELECT * FROM ducknng_http_text(
     201,
     (SELECT method || '' '' || path || ''?'' || coalesce(query_string, '''') ||
             '' x='' || coalesce(ducknng_http_query_param(''x''), '''') ||
             '' '' || coalesce(content_type, '''') ||
             '' '' || coalesce(body_text, '''')
      FROM ducknng_http_request(), ducknng_http_request_body()))'
);

SELECT ducknng_register_http_route_pattern(
  'http_route_demo', 'GET', 'template',
  '/tenant/{tenant_id}/items/{item_id}',
  'SELECT * FROM ducknng_http_text(
     202,
     (SELECT route_match_kind || '' '' ||
             ducknng_http_path_param(''tenant_id'') || '':'' ||
             ducknng_http_path_param(''item_id'')
      FROM ducknng_http_request()))'
);
```

#### Send test requests

```{duckdb collapse=TRUE, comment=""}
SELECT ok, status = 200, body_text
FROM ducknng_ncurl('http://127.0.0.1:18445/healthz', 'GET', NULL, NULL, 2000, 0::UBIGINT);

SELECT ok, status = 201, body_text
FROM ducknng_ncurl(
  'http://127.0.0.1:18445/echo?x=1', 'POST',
  '[{"name":"Content-Type","value":"text/plain"}]',
  'hello'::BLOB, 2000, 0::UBIGINT
);

SELECT ok, status = 202, body_text
FROM ducknng_ncurl(
  'http://127.0.0.1:18445/tenant/alice/items/42', 'GET', NULL, NULL, 2000, 0::UBIGINT
);
```

#### List routes and stop the server

```{duckdb collapse=TRUE, comment=""}
SELECT service_name, method, match_kind, path
FROM ducknng_list_http_routes()
WHERE service_name = 'http_route_demo'
ORDER BY match_kind, path;

SELECT ducknng_stop_server('http_route_demo');
```

By default these route handlers run on the backward-compatible `shared_serialized_connection` lane, but services can switch before traffic to `service_serialized_connection` or `request_connection` with `ducknng_set_service_execution_model(...)`. Those models use pre-opened DuckDB execution-pool connections from the same database handle, which improves same-runtime gateway composition but means handlers should depend on catalog-visible objects rather than temp tables or temp macros from the init connection. Prefix and template routes stay low-level on purpose: the request context remains explicit, while `ducknng_http_header(...)`, `ducknng_http_query_param(...)`, `ducknng_http_cookie(...)`, and `ducknng_http_path_param(...)` remove the repetitive JSON/string parsing from common handlers. The response helpers such as `ducknng_http_text(...)` and `ducknng_http_json(...)` only build the existing one-row route response shape. That keeps the route framework additive instead of turning it into a second RPC namespace. They are not a safe excuse to expose arbitrary SQL to the public internet.

A public subscriber gateway is a good fit for this layer, but it should use separate backend DuckDB processes or runtime boundaries. The main README still stays single-session, so the multi-process walkthrough lives separately: `docs/subscriber_gateway_demo.md` explains the topology, `demo/subscriber_gateway.py` is the live helper, `make subscriber_gateway_demo` runs the end-to-end check, and `make subscriber_gateway_rdm` renders a dedicated `demo/subscriber_gateway.Rmd` walkthrough with hidden worker setup and live HTTP requests.

### Launch raw socket send/recv airos and inspect send status explicitly

#### Open and connect the socket pair

```{duckdb collapse=TRUE, comment=""}
SET VARIABLE pair_a = (ducknng_open_socket('pair')).socket_id;
SET VARIABLE pair_listen_ok = (ducknng_listen_socket(
  getvariable('pair_a')::UBIGINT,
  'ipc:///tmp/ducknng_sql_pair_aio_demo.ipc',
  134217728, 0::UBIGINT
)).ok;
SET VARIABLE pair_b = (ducknng_open_socket('pair')).socket_id;
SET VARIABLE pair_dial_ok = (ducknng_dial_socket(
  getvariable('pair_b')::UBIGINT,
  'ipc:///tmp/ducknng_sql_pair_aio_demo.ipc',
  1000, 0::UBIGINT
)).ok;
```

#### Launch async send and receive, then collect results

```{duckdb collapse=TRUE, comment=""}
CREATE TEMP TABLE pair_recv AS
SELECT ducknng_recv_socket_raw_aio(getvariable('pair_a')::UBIGINT, 1000) AS recv_aio;

CREATE TEMP TABLE pair_send AS
SELECT ducknng_send_socket_raw_aio(
  getvariable('pair_b')::UBIGINT,
  from_hex('6173796e632072657175657374'),
  1000
) AS send_aio;

-- Send-only airos succeed with frame = NULL.
SELECT aio_id, ok, frame IS NULL AS no_frame
FROM ducknng_aio_collect((SELECT list_value(send_aio) FROM pair_send), 1000);

SELECT aio_id, ok, hex(frame) = '6173796E632072657175657374' AS got_payload
FROM ducknng_aio_collect((SELECT list_value(recv_aio) FROM pair_recv), 1000);

SELECT kind, state, phase, send_done, send_ok, recv_done, recv_ok IS NULL AS recv_ok_is_null
FROM ducknng_aio_status((SELECT send_aio FROM pair_send));
```

#### Clean up

```{duckdb collapse=TRUE, comment=""}
SELECT ducknng_aio_drop((SELECT send_aio FROM pair_send)) AS dropped_send,
       ducknng_aio_drop((SELECT recv_aio FROM pair_recv)) AS dropped_recv;
DROP TABLE pair_send;
DROP TABLE pair_recv;
SELECT (ducknng_close_socket(getvariable('pair_b')::UBIGINT)).ok AS closed_b,
       (ducknng_close_socket(getvariable('pair_a')::UBIGINT)).ok AS closed_a;
```

### Push one raw message through `push` / `pull`

#### Open and connect

```{duckdb collapse=TRUE, comment=""}
SET VARIABLE pull_socket = (ducknng_open_socket('pull')).socket_id;
SET VARIABLE pull_listen_ok = (ducknng_listen_socket(
  getvariable('pull_socket')::UBIGINT,
  'ipc:///tmp/ducknng_sql_pushpull_demo.ipc',
  134217728, 0::UBIGINT
)).ok;
SET VARIABLE push_socket = (ducknng_open_socket('push')).socket_id;
SET VARIABLE push_dial_ok = (ducknng_dial_socket(
  getvariable('push_socket')::UBIGINT,
  'ipc:///tmp/ducknng_sql_pushpull_demo.ipc',
  1000, 0::UBIGINT
)).ok;
```

#### Send and receive

```{duckdb collapse=TRUE, comment=""}
CREATE TEMP TABLE pushpull_recv AS
SELECT ducknng_recv_socket_raw_aio(getvariable('pull_socket')::UBIGINT, 1000) AS recv_aio;

SET VARIABLE pushpull_sent = (ducknng_send_socket_raw(
  getvariable('push_socket')::UBIGINT, from_hex('707573682d70756c6c'), 1000
)).ok;

SELECT getvariable('pushpull_sent')::BOOLEAN AS sent,
       ok, hex(frame) = '707573682D70756C6C' AS got_payload
FROM ducknng_aio_collect((SELECT list_value(recv_aio) FROM pushpull_recv), 1000);
```

#### Clean up

```{duckdb collapse=TRUE, comment=""}
SELECT ducknng_aio_drop((SELECT recv_aio FROM pushpull_recv)) AS dropped;
DROP TABLE pushpull_recv;
SELECT (ducknng_close_socket(getvariable('push_socket')::UBIGINT)).ok AS closed_push,
       (ducknng_close_socket(getvariable('pull_socket')::UBIGINT)).ok AS closed_pull;
```

### Publish one raw message through `pub` / `sub`

#### Open, subscribe, and connect

```{duckdb collapse=TRUE, comment=""}
SET VARIABLE pub_socket = (ducknng_open_socket('pub')).socket_id;
SET VARIABLE pub_listen_ok = (ducknng_listen_socket(
  getvariable('pub_socket')::UBIGINT,
  'ipc:///tmp/ducknng_sql_pubsub_demo.ipc',
  134217728, 0::UBIGINT
)).ok;
SET VARIABLE sub_socket = (ducknng_open_socket('sub')).socket_id;
SET VARIABLE sub_subscribe_ok = (ducknng_subscribe_socket(
  getvariable('sub_socket')::UBIGINT, from_hex('')
)).ok;
SET VARIABLE sub_dial_ok = (ducknng_dial_socket(
  getvariable('sub_socket')::UBIGINT,
  'ipc:///tmp/ducknng_sql_pubsub_demo.ipc',
  1000, 0::UBIGINT
)).ok;
```

#### Publish and receive

```{duckdb collapse=TRUE, comment=""}
CREATE TEMP TABLE pubsub_recv AS
SELECT ducknng_recv_socket_raw_aio(getvariable('sub_socket')::UBIGINT, 1000) AS recv_aio;

SET VARIABLE pubsub_sent = (ducknng_send_socket_raw(
  getvariable('pub_socket')::UBIGINT, from_hex('7075622d737562'), 1000
)).ok;

SELECT getvariable('pubsub_sent')::BOOLEAN AS sent,
       ok, hex(frame) = '7075622D737562' AS got_payload
FROM ducknng_aio_collect((SELECT list_value(recv_aio) FROM pubsub_recv), 1000);
```

#### Clean up

```{duckdb collapse=TRUE, comment=""}
SELECT ducknng_aio_drop((SELECT recv_aio FROM pubsub_recv)) AS dropped;
DROP TABLE pubsub_recv;
SELECT (ducknng_close_socket(getvariable('sub_socket')::UBIGINT)).ok AS closed_sub,
       (ducknng_close_socket(getvariable('pub_socket')::UBIGINT)).ok AS closed_pub;
```

### Exchange one survey and one response through `surveyor` / `respondent`

#### Open respondent listener and surveyor peer

```{duckdb collapse=TRUE, comment=""}
SET VARIABLE respondent_socket = (ducknng_open_socket('respondent')).socket_id;
SET VARIABLE respondent_listen_ok = (ducknng_listen_socket(
  getvariable('respondent_socket')::UBIGINT,
  'ipc:///tmp/ducknng_sql_survey_demo.ipc',
  134217728, 0::UBIGINT
)).ok;
SET VARIABLE surveyor_socket = (ducknng_open_socket('surveyor')).socket_id;
SET VARIABLE surveyor_dial_ok = (ducknng_dial_socket(
  getvariable('surveyor_socket')::UBIGINT,
  'ipc:///tmp/ducknng_sql_survey_demo.ipc',
  1000, 0::UBIGINT
)).ok;
```

#### Exchange the survey and the response

```{duckdb collapse=TRUE, comment=""}
CREATE TEMP TABLE respondent_recv AS
SELECT ducknng_recv_socket_raw_aio(getvariable('respondent_socket')::UBIGINT, 1000) AS recv_aio;

SET VARIABLE survey_sent = (ducknng_send_socket_raw(
  getvariable('surveyor_socket')::UBIGINT, from_hex('737572766579'), 1000
)).ok;

SELECT getvariable('survey_sent')::BOOLEAN AS sent_survey,
       ok, hex(frame) = '737572766579' AS got_survey
FROM ducknng_aio_collect((SELECT list_value(recv_aio) FROM respondent_recv), 1000);

CREATE TEMP TABLE surveyor_recv AS
SELECT ducknng_recv_socket_raw_aio(getvariable('surveyor_socket')::UBIGINT, 1000) AS recv_aio;

SET VARIABLE response_sent = (ducknng_send_socket_raw(
  getvariable('respondent_socket')::UBIGINT, from_hex('726573706f6e7365'), 1000
)).ok;

SELECT getvariable('response_sent')::BOOLEAN AS sent_response,
       ok, hex(frame) = '726573706F6E7365' AS got_response
FROM ducknng_aio_collect((SELECT list_value(recv_aio) FROM surveyor_recv), 1000);
```

#### Clean up

```{duckdb collapse=TRUE, comment=""}
SELECT ducknng_aio_drop((SELECT recv_aio FROM respondent_recv)) AS dropped_respondent,
       ducknng_aio_drop((SELECT recv_aio FROM surveyor_recv)) AS dropped_surveyor;
DROP TABLE respondent_recv;
DROP TABLE surveyor_recv;
SELECT (ducknng_close_socket(getvariable('surveyor_socket')::UBIGINT)).ok AS closed_surveyor,
       (ducknng_close_socket(getvariable('respondent_socket')::UBIGINT)).ok AS closed_respondent;
```

### Broadcast one raw message through a three-peer `bus` mesh

Every peer in a bus sends to and receives from every other peer, so a single send from peer A is delivered to both B and C.

#### Open three bus peers and connect

```{duckdb collapse=TRUE, comment=""}
SET VARIABLE bus_a = (ducknng_open_socket('bus')).socket_id;
SET VARIABLE bus_b = (ducknng_open_socket('bus')).socket_id;
SET VARIABLE bus_c = (ducknng_open_socket('bus')).socket_id;

-- A listens; B and C dial A for a fully connected mesh.
SET VARIABLE bus_a_listen_ok = (ducknng_listen_socket(
  getvariable('bus_a')::UBIGINT,
  'ipc:///tmp/ducknng_sql_bus_demo.ipc',
  134217728, 0::UBIGINT
)).ok;
SET VARIABLE bus_b_dial_ok = (ducknng_dial_socket(
  getvariable('bus_b')::UBIGINT,
  'ipc:///tmp/ducknng_sql_bus_demo.ipc',
  1000, 0::UBIGINT
)).ok;
SET VARIABLE bus_c_dial_ok = (ducknng_dial_socket(
  getvariable('bus_c')::UBIGINT,
  'ipc:///tmp/ducknng_sql_bus_demo.ipc',
  1000, 0::UBIGINT
)).ok;
```

#### Arm receivers, broadcast from A, and collect at B and C

```{duckdb collapse=TRUE, comment=""}
CREATE TEMP TABLE bus_recv_b AS
SELECT ducknng_recv_socket_raw_aio(getvariable('bus_b')::UBIGINT, 1000) AS recv_aio;
CREATE TEMP TABLE bus_recv_c AS
SELECT ducknng_recv_socket_raw_aio(getvariable('bus_c')::UBIGINT, 1000) AS recv_aio;

SET VARIABLE bus_sent = (ducknng_send_socket_raw(
  getvariable('bus_a')::UBIGINT, from_hex('6275732d62726f616463617374'), 1000
)).ok;

SELECT getvariable('bus_sent')::BOOLEAN AS sent,
       ok, hex(frame) = '6275732D62726F616463617374' AS got_payload
FROM ducknng_aio_collect((SELECT list_value(recv_aio) FROM bus_recv_b), 1000);

SELECT ok, hex(frame) = '6275732D62726F616463617374' AS got_payload
FROM ducknng_aio_collect((SELECT list_value(recv_aio) FROM bus_recv_c), 1000);
```

#### Clean up

```{duckdb collapse=TRUE, comment=""}
SELECT ducknng_aio_drop((SELECT recv_aio FROM bus_recv_b)) AS dropped_b,
       ducknng_aio_drop((SELECT recv_aio FROM bus_recv_c)) AS dropped_c;
DROP TABLE bus_recv_b;
DROP TABLE bus_recv_c;
SELECT (ducknng_close_socket(getvariable('bus_c')::UBIGINT)).ok AS closed_c,
       (ducknng_close_socket(getvariable('bus_b')::UBIGINT)).ok AS closed_b,
       (ducknng_close_socket(getvariable('bus_a')::UBIGINT)).ok AS closed_a;
```

### Launch raw requests asynchronously and collect the reply frames later

The stable async contract is raw-result-first. Framed RPC aio helpers collect raw reply frames through `ducknng_aio_collect(...)`, low-level HTTP aio helpers collect HTTP-shaped rows through `ducknng_ncurl_aio_collect(...)`, and callers explicitly decode frames or bodies afterward. Expected aio launch failures, such as unsupported URL schemes, invalid TLS handles, or missing socket handles, return immediate terminal error handles so the caller can inspect `ducknng_aio_status(...)` or collect the error row without turning the launch itself into a DuckDB exception. `ducknng_aio_wait(...)` is the wait-without-consuming primitive for lifecycle code that needs to inspect or drop a terminal handle later. `ducknng_aio_collect_decoded(...)` projects the decoded envelope columns directly through the same low-level frame scalar accessors. Aio handles represent one pending operation, not a background job or streaming protocol.

#### Start the listener and launch two parallel requests

```{duckdb collapse=TRUE, comment=""}
SELECT ducknng_start_server(
  'sql_aio_demo', 'ipc:///tmp/ducknng_sql_aio_demo.ipc',
  1, 134217728, 300000, 0
);

-- timeout_ms bounds the pending network op; wait_ms in collect() is separate.
CREATE TEMP TABLE aio_demo AS
SELECT
  ducknng_request_raw_aio(
    'ipc:///tmp/ducknng_sql_aio_demo.ipc',
    from_hex('01000000000000000000000000000000000000000000'),
    1000, 0::UBIGINT
  ) AS aio1,
  ducknng_request_raw_aio(
    'ipc:///tmp/ducknng_sql_aio_demo.ipc',
    from_hex('01000000000000000000000000000000000000000000'),
    1000, 0::UBIGINT
  ) AS aio2;

SELECT aio1 > 0 AS aio1_started, aio2 > aio1 AS aio2_started_after_aio1
FROM aio_demo;
```

#### Wait, collect, and inspect terminal handles

```{duckdb collapse=TRUE, comment=""}
-- Wait without consuming — status/drop logic can still run after this.
SELECT ducknng_aio_wait((SELECT list_value(aio1, aio2) FROM aio_demo), 1000) AS any_ready;

SELECT aio_id, ok, octet_length(frame) > 0 AS has_frame
FROM ducknng_aio_collect((SELECT list_value(aio1, aio2) FROM aio_demo), 0)
ORDER BY aio_id;

-- A collected aio stays terminal (ready = true) until explicitly dropped.
SELECT ducknng_aio_ready(aio1) AS aio1_ready, ducknng_aio_ready(aio2) AS aio2_ready
FROM aio_demo;
```

#### Drop handles and stop the server

```{duckdb collapse=TRUE, comment=""}
SELECT ducknng_aio_drop(aio1) AND ducknng_aio_drop(aio2) AS dropped
FROM aio_demo;
DROP TABLE aio_demo;
SELECT ducknng_stop_server('sql_aio_demo');
```

### Launch unary RPC calls asynchronously and decode the replies later

These helpers build the manifest and exec request frames for you and sit above the same `ducknng_request_raw_aio(...)` substrate. Decoding remains explicit.

#### Start the server

`exec` is already in the global registry from the earlier example. A fresh server inherits the current registry automatically.

```{duckdb collapse=TRUE, comment=""}
SELECT ducknng_start_server(
  'sql_rpc_aio_demo', 'ipc:///tmp/ducknng_sql_rpc_aio_demo.ipc',
  1, 134217728, 300000, 0
);
```

#### Launch manifest and exec aio calls

```{duckdb collapse=TRUE, comment=""}
SET VARIABLE manifest_aio = ducknng_get_rpc_manifest_raw_aio(
  'ipc:///tmp/ducknng_sql_rpc_aio_demo.ipc', 1000, 0::UBIGINT
);
SET VARIABLE exec_aio = ducknng_run_rpc_raw_aio(
  'ipc:///tmp/ducknng_sql_rpc_aio_demo.ipc',
  'CREATE TABLE IF NOT EXISTS rpc_aio_demo_t(i INTEGER)',
  1000, 0::UBIGINT
);

SELECT getvariable('manifest_aio') > 0 AS manifest_aio_started,
       getvariable('exec_aio') > getvariable('manifest_aio') AS exec_aio_after_manifest;
```

#### Collect and decode the frames

```{duckdb collapse=TRUE, comment=""}
CREATE TEMP TABLE rpc_aio_collect AS
SELECT *
FROM ducknng_aio_collect(
  list_value(getvariable('manifest_aio'), getvariable('exec_aio')), 1000
);

SET VARIABLE manifest_frame = (
  SELECT frame FROM rpc_aio_collect WHERE aio_id = getvariable('manifest_aio')
);
SET VARIABLE exec_frame = (
  SELECT frame FROM rpc_aio_collect WHERE aio_id = getvariable('exec_aio')
);

SELECT ok, type_name, name,
       position('"name":"exec"' IN payload_text) > 0 AS has_exec
FROM ducknng_decode_frame(getvariable('manifest_frame'));

SELECT ok, type_name, name
FROM ducknng_decode_frame(getvariable('exec_frame'));
```

#### Convenience decoded collection wrapper

```{duckdb collapse=TRUE, comment=""}
SET VARIABLE manifest_aio_decoded = ducknng_get_rpc_manifest_raw_aio(
  'ipc:///tmp/ducknng_sql_rpc_aio_demo.ipc', 1000, 0::UBIGINT
);

SELECT aio_id, ok, frame_ok, type_name, name
FROM ducknng_aio_collect_decoded(list_value(getvariable('manifest_aio_decoded')), 1000);
```

#### Clean up

```{duckdb collapse=TRUE, comment=""}
SELECT ducknng_aio_drop(getvariable('manifest_aio')) AND
       ducknng_aio_drop(getvariable('exec_aio')) AND
       ducknng_aio_drop(getvariable('manifest_aio_decoded')) AS dropped;
DROP TABLE rpc_aio_collect;
SELECT ducknng_stop_server('sql_rpc_aio_demo');
```

### Open, fetch, and close a query session explicitly

#### Start the server and open a session

```{duckdb collapse=TRUE, comment=""}
SELECT ducknng_start_server(
  'sql_session_demo', 'ipc:///tmp/ducknng_sql_session_demo.ipc',
  1, 134217728, 300000, 0
);

-- batch_rows = 0 and batch_bytes = 0 use server defaults.
-- Keep both session_id and session_token: the token is the bearer capability
-- required by fetch, close, and cancel.
CREATE TEMP TABLE session_open AS
SELECT *
FROM ducknng_open_query(
  'ipc:///tmp/ducknng_sql_session_demo.ipc',
  'SELECT 1 AS id UNION ALL SELECT 2 AS id ORDER BY id',
  0::UBIGINT, 0::UBIGINT, 0::UBIGINT
);

SET VARIABLE session_id    = (SELECT session_id    FROM session_open);
SET VARIABLE session_token = (SELECT session_token FROM session_open);

SELECT getvariable('session_id') AS opened_session_id,
       length(getvariable('session_token')::VARCHAR) AS session_token_chars,
       idle_timeout_ms AS effective_idle_timeout_ms
FROM session_open;
```

#### Fetch rows and check the exhausted state

```{duckdb collapse=TRUE, comment=""}
-- Table helper decodes the first batch directly as rows.
SELECT *
FROM ducknng_fetch_query_table(
  'ipc:///tmp/ducknng_sql_session_demo.ipc',
  getvariable('session_id')::UBIGINT,
  getvariable('session_token')::VARCHAR,
  0::UBIGINT, 0::UBIGINT, 0::UBIGINT
);

-- After the table helper consumes the batch, the next fetch returns the
-- exhausted control reply (end_of_stream = true, no payload).
SELECT ok, session_id, state, payload IS NULL AS no_payload, end_of_stream
FROM ducknng_fetch_query(
  'ipc:///tmp/ducknng_sql_session_demo.ipc',
  getvariable('session_id')::UBIGINT,
  getvariable('session_token')::VARCHAR,
  0::UBIGINT, 0::UBIGINT, 0::UBIGINT
);
```

#### Close the session and stop the server

```{duckdb collapse=TRUE, comment=""}
SELECT ok, session_id, state
FROM ducknng_close_query(
  'ipc:///tmp/ducknng_sql_session_demo.ipc',
  getvariable('session_id')::UBIGINT,
  getvariable('session_token')::VARCHAR,
  0::UBIGINT
);

DROP TABLE session_open;
SELECT ducknng_stop_server('sql_session_demo');
```

### Drive the same session lifecycle through raw frames

#### Start the server and open a raw session

```{duckdb collapse=TRUE, comment=""}
SELECT ducknng_start_server(
  'sql_session_raw_demo', 'ipc:///tmp/ducknng_sql_session_raw_demo.ipc',
  1, 134217728, 300000, 0
);

SET VARIABLE raw_session_open_frame = ducknng_open_query_raw(
  'ipc:///tmp/ducknng_sql_session_raw_demo.ipc',
  'SELECT 1 AS id UNION ALL SELECT 2 AS id ORDER BY id',
  0::UBIGINT, 0::UBIGINT, 0::UBIGINT
);

SELECT ducknng_frame_error_text(getvariable('raw_session_open_frame')::BLOB) AS open_error,
       position('"session_id":' IN ducknng_frame_payload_text(getvariable('raw_session_open_frame')::BLOB)) > 0 AS has_session_id;

SET VARIABLE raw_session_id = (
  SELECT json_extract(
    ducknng_frame_payload_text(getvariable('raw_session_open_frame')::BLOB)::JSON,
    '$.session_id'
  )::UBIGINT
);
SET VARIABLE raw_session_token = (
  SELECT json_extract_string(
    ducknng_frame_payload_text(getvariable('raw_session_open_frame')::BLOB)::JSON,
    '$.session_token'
  )
);
```

#### Fetch a batch and decode the Arrow payload

```{duckdb collapse=TRUE, comment=""}
SET VARIABLE raw_session_fetch_frame = ducknng_fetch_query_raw(
  'ipc:///tmp/ducknng_sql_session_raw_demo.ipc',
  getvariable('raw_session_id')::UBIGINT,
  getvariable('raw_session_token')::VARCHAR,
  0::UBIGINT, 0::UBIGINT, 0::UBIGINT
);

SELECT ducknng_frame_version(getvariable('raw_session_fetch_frame')::BLOB)      AS version,
       ducknng_frame_type_name(getvariable('raw_session_fetch_frame')::BLOB)    AS type_name,
       ducknng_frame_name(getvariable('raw_session_fetch_frame')::BLOB)         AS name,
       ducknng_frame_end_of_stream(getvariable('raw_session_fetch_frame')::BLOB) AS end_of_stream,
       ducknng_frame_flags(getvariable('raw_session_fetch_frame')::BLOB) > 0    AS has_flags;

SELECT *
FROM ducknng_parse_body(
  ducknng_frame_payload(getvariable('raw_session_fetch_frame')::BLOB),
  'application/vnd.apache.arrow.stream'
);
```

#### Close the raw session and stop the server

```{duckdb collapse=TRUE, comment=""}
SET VARIABLE raw_session_close_frame = ducknng_close_query_raw(
  'ipc:///tmp/ducknng_sql_session_raw_demo.ipc',
  getvariable('raw_session_id')::UBIGINT,
  getvariable('raw_session_token')::VARCHAR,
  0::UBIGINT
);

SELECT position('"state":"closed"' IN ducknng_frame_payload_text(
  getvariable('raw_session_close_frame')::BLOB
)) > 0 AS is_closed;

SELECT ducknng_stop_server('sql_session_raw_demo');
```

### Launch the session lifecycle asynchronously and collect raw frames

#### Start the server

```{duckdb collapse=TRUE, comment=""}
SELECT ducknng_start_server(
  'sql_session_aio_demo', 'ipc:///tmp/ducknng_sql_session_aio_demo.ipc',
  1, 134217728, 300000, 0
);
```

#### Open session asynchronously

```{duckdb collapse=TRUE, comment=""}
SET VARIABLE session_open_aio = ducknng_open_query_raw_aio(
  'ipc:///tmp/ducknng_sql_session_aio_demo.ipc',
  'SELECT 1 AS id UNION ALL SELECT 2 AS id ORDER BY id',
  0::UBIGINT, 0::UBIGINT, 1000, 0::UBIGINT
);

CREATE TEMP TABLE session_open_aio_collect AS
SELECT * FROM ducknng_aio_collect(list_value(getvariable('session_open_aio')::UBIGINT), 1000);

SET VARIABLE session_open_frame = (SELECT frame FROM session_open_aio_collect);

SELECT ok, type_name, name,
       position('"session_id":' IN payload_text) > 0 AS has_session_id
FROM ducknng_decode_frame(getvariable('session_open_frame')::BLOB);

SET VARIABLE session_aio_id = (
  SELECT json_extract(payload_text::JSON, '$.session_id')::UBIGINT
  FROM ducknng_decode_frame(getvariable('session_open_frame')::BLOB)
);
SET VARIABLE session_aio_token = (
  SELECT json_extract_string(payload_text::JSON, '$.session_token')
  FROM ducknng_decode_frame(getvariable('session_open_frame')::BLOB)
);

SELECT ducknng_aio_drop(getvariable('session_open_aio')::UBIGINT) AS dropped_open_aio;
DROP TABLE session_open_aio_collect;
```

#### Fetch asynchronously

```{duckdb collapse=TRUE, comment=""}
SET VARIABLE session_fetch_aio = ducknng_fetch_query_raw_aio(
  'ipc:///tmp/ducknng_sql_session_aio_demo.ipc',
  getvariable('session_aio_id')::UBIGINT,
  getvariable('session_aio_token')::VARCHAR,
  0::UBIGINT, 0::UBIGINT, 1000, 0::UBIGINT
);

CREATE TEMP TABLE session_fetch_aio_collect AS
SELECT * FROM ducknng_aio_collect(list_value(getvariable('session_fetch_aio')::UBIGINT), 1000);

SET VARIABLE session_fetch_frame = (SELECT frame FROM session_fetch_aio_collect);

SELECT ok, type_name, name, octet_length(payload) > 0 AS has_payload
FROM ducknng_decode_frame(getvariable('session_fetch_frame')::BLOB);

SELECT ducknng_aio_drop(getvariable('session_fetch_aio')::UBIGINT) AS dropped_fetch_aio;
DROP TABLE session_fetch_aio_collect;
```

#### Close asynchronously

```{duckdb collapse=TRUE, comment=""}
SET VARIABLE session_close_aio = ducknng_close_query_raw_aio(
  'ipc:///tmp/ducknng_sql_session_aio_demo.ipc',
  getvariable('session_aio_id')::UBIGINT,
  getvariable('session_aio_token')::VARCHAR,
  1000, 0::UBIGINT
);

CREATE TEMP TABLE session_close_aio_collect AS
SELECT * FROM ducknng_aio_collect(list_value(getvariable('session_close_aio')::UBIGINT), 1000);

SET VARIABLE session_close_frame = (SELECT frame FROM session_close_aio_collect);

SELECT ok, type_name, name,
       position('"state":"closed"' IN payload_text) > 0 AS is_closed
FROM ducknng_decode_frame(getvariable('session_close_frame')::BLOB);

SELECT ducknng_aio_drop(getvariable('session_close_aio')::UBIGINT) AS dropped_close_aio;
DROP TABLE session_close_aio_collect;
```

#### Cancel asynchronously

```{duckdb collapse=TRUE, comment=""}
CREATE TEMP TABLE session_cancel_open AS
SELECT * FROM ducknng_open_query(
  'ipc:///tmp/ducknng_sql_session_aio_demo.ipc',
  'SELECT 99 AS id', 0::UBIGINT, 0::UBIGINT, 0::UBIGINT
);

SET VARIABLE session_cancel_id    = (SELECT session_id    FROM session_cancel_open);
SET VARIABLE session_cancel_token = (SELECT session_token FROM session_cancel_open);

SET VARIABLE session_cancel_aio = ducknng_cancel_query_raw_aio(
  'ipc:///tmp/ducknng_sql_session_aio_demo.ipc',
  getvariable('session_cancel_id')::UBIGINT,
  getvariable('session_cancel_token')::VARCHAR,
  1000, 0::UBIGINT
);

CREATE TEMP TABLE session_cancel_aio_collect AS
SELECT * FROM ducknng_aio_collect(list_value(getvariable('session_cancel_aio')::UBIGINT), 1000);

SET VARIABLE session_cancel_frame = (SELECT frame FROM session_cancel_aio_collect);

SELECT ok, type_name, name,
       position('"state":"cancelled"' IN payload_text) > 0 AS is_cancelled
FROM ducknng_decode_frame(getvariable('session_cancel_frame')::BLOB);

SELECT ducknng_aio_drop(getvariable('session_cancel_aio')::UBIGINT) AS dropped_cancel_aio;
DROP TABLE session_cancel_open;
DROP TABLE session_cancel_aio_collect;
```

#### Stop the server

```{duckdb collapse=TRUE, comment=""}
SELECT ducknng_stop_server('sql_session_aio_demo');
```

### `tls+tcp://` with a self-signed development TLS config

```{duckdb collapse=TRUE, comment=""}
-- Generate a self-signed loopback certificate and keep it as a runtime TLS handle.
SET VARIABLE tls_self_id = ducknng_self_signed_tls_config('127.0.0.1', 365, 0);

-- Start a TLS listener using the generated config handle.
SELECT ducknng_start_server(
  'tls_demo_self',               -- service name
  'tls+tcp://127.0.0.1:45453',   -- listen URL
  1,                             -- REP contexts
  134217728,                     -- recv_max_bytes
  300000,                        -- session_idle_ms
  getvariable('tls_self_id')::UBIGINT  -- tls_config_id returned above
);

-- Send the built-in manifest request frame over TLS and decode the reply.
SELECT ok, type_name, name, position('"name":"exec"' IN payload_text) > 0
FROM ducknng_decode_frame(
  ducknng_request_raw(
    'tls+tcp://127.0.0.1:45453',                -- url
    from_hex('01000000000000000000000000000000000000000000'), -- manifest request frame
    1000,                                       -- timeout_ms
    getvariable('tls_self_id')::UBIGINT         -- tls_config_id
  )
);

-- Clean up the TLS demo server and config handle.
SELECT ducknng_stop_server('tls_demo_self');
SELECT ducknng_drop_tls_config(getvariable('tls_self_id')::UBIGINT);
```

For mTLS, create the same kind of TLS handle with `auth_mode = 2`. On listeners this requires the peer to present a certificate trusted by the configured CA; on clients it requires server verification. The dispatcher derives the current caller identity from the first verified peer certificate SAN as `tls:san:<value>`, falling back to the common name as `tls:cn:<common-name>`. Sessions opened over mTLS are still controlled by `session_token`, but they are also bound to that verified peer identity. `ducknng_set_tls_peer_allowlist(...)` sets an exact identity allowlist copied into future services, and `ducknng_set_service_peer_allowlist(...)` changes the allowlist dynamically for a running service. NNG listeners use NNG's `NNG_PIPE_EV_ADD_PRE` pipe notification to close non-admitted new pipes before they are added to the socket; HTTP/HTTPS returns HTTP `403` before RPC dispatch. `ducknng_list_servers()` exposes `execution_model`, `active_pipes`, `max_active_pipes`, `inflight_requests`, `max_inflight_requests`, `max_sessions_per_peer_identity`, `tls_enabled`, `tls_auth_mode`, `peer_identity_required`, `peer_allowlist_active`, `ip_allowlist_active`, `sql_authorizer_active`, and the corresponding counts so deployments can distinguish the current execution-lane contract, current NNG membership, TLS without client verification, mTLS, allowlisted mTLS, IP-gated services, and services with SQL authorization callbacks. Individual registry-backed methods can also require verified peer identity; for example, `ducknng_set_method_auth('manifest', true)` protects manifest discovery without unregistering the method.

The built-in mTLS, peer-identity allowlist, and IP/CIDR allowlist checks are the fast C path for common denials. They run before the flexible SQL callback and, for NNG transports, can reject new pipes at `NNG_PIPE_EV_ADD_PRE` without entering DuckDB. Basic service resource limits also start in C: `ducknng_set_service_limits(name, max_open_sessions)` caps concurrently open query sessions for a service, the three-argument form also caps simultaneously active NNG protocol-socket pipes at `ADD_PRE`, and the four-argument form also caps concurrent in-flight requests before SQL authorizers and RPC dispatch, and the five-argument form also caps concurrently open query sessions per verified peer identity; `0` means unlimited for any cap. Those are the stable built-in quota owners today: the service itself and, when present, the verified peer identity. The optional SQL-authorizer `principal` column is deployment policy/audit metadata, not yet durable session ownership metadata for built-in quotas. For policy that depends on tables, tenants, method names, headers, or deployment-specific rules, install a service-level SQL authorizer with `ducknng_set_service_authorizer(name, authorizer_sql)`. The callback sees one row from `ducknng_auth_context()` while it runs and must return exactly one row with a Boolean `allow` column; optional columns are `status`, `reason`, `principal`, `claims_json`, and `cache_ttl_ms`. `NULL` or an empty string clears the SQL authorizer. This callback is intentionally evaluated at the request/dispatch boundary, not inside NNG's low-level pipe callback. It runs through the configured service-owned DuckDB execution model, so keep it short and side-effect-light: prefer table/view lookups, avoid recursive `ducknng_*` client calls to the same service, avoid stopping the service from its own callback, and do not use it for long-running work. That limitation avoids deadlocks while keeping one uniform policy interface for NNG, HTTP/HTTPS framed RPC, and HTTP routes.

### `tls+tcp://` from file-backed certificate material

```{duckdb collapse=TRUE, comment=""}
-- Register a file-backed TLS config using committed loopback test certificates.
SET VARIABLE tls_files_id = ducknng_tls_config_from_files(
  'test/certs/loopback-cert-key.pem', -- cert_key_file
  'test/certs/loopback-ca.pem',       -- ca_file
  NULL,                               -- password
  0                                   -- auth_mode
);

-- Start a TLS listener with that file-backed config.
SELECT ducknng_start_server(
  'tls_demo_files',              -- service name
  'tls+tcp://127.0.0.1:45454',   -- listen URL
  1,                             -- REP contexts
  134217728,                     -- recv_max_bytes
  300000,                        -- session_idle_ms
  getvariable('tls_files_id')::UBIGINT  -- tls_config_id returned above
);

-- Inspect the manifest reply over TLS using the same config handle on the client side.
SELECT ok, type_name, name, position('"name":"exec"' IN payload_text) > 0
FROM ducknng_decode_frame(
  ducknng_request_raw(
    'tls+tcp://127.0.0.1:45454',                -- url
    from_hex('01000000000000000000000000000000000000000000'), -- manifest request frame
    1000,                                       -- timeout_ms
    getvariable('tls_files_id')::UBIGINT        -- tls_config_id
  )
);

-- Clean up the file-backed TLS demo server and config handle.
SELECT ducknng_stop_server('tls_demo_files');
SELECT ducknng_drop_tls_config(getvariable('tls_files_id')::UBIGINT);
```

### `ws://` and `wss://` as NNG transports

`ws://` and `wss://` are NNG transport schemes, not the HTTP carrier described in `docs/http.md`. They use `ducknng_start_server(...)`, the same framed manifest/RPC/session helpers, and the same TLS handle model as the other NNG transports.

```{duckdb collapse=TRUE, comment=""}
-- Plain WebSocket transport through the NNG service layer.
SELECT ducknng_start_server(
  'ws_demo',
  'ws://127.0.0.1:45455/_ducknng',
  1,
  134217728,
  300000,
  0::UBIGINT
);

SELECT ok, type_name, name
FROM ducknng_decode_frame(
  ducknng_request_raw(
    'ws://127.0.0.1:45455/_ducknng',
    from_hex('01000000000000000000000000000000000000000000'),
    1000,
    0::UBIGINT
  )
);

SELECT ducknng_stop_server('ws_demo');

-- Secure WebSocket transport uses the same reusable TLS handle model.
SET VARIABLE tls_wss_id = ducknng_self_signed_tls_config('127.0.0.1', 365, 0);

SELECT ducknng_start_server(
  'wss_demo',
  'wss://127.0.0.1:45456/_ducknng',
  1,
  134217728,
  300000,
  getvariable('tls_wss_id')::UBIGINT
);

SELECT ok, type_name, name
FROM ducknng_decode_frame(
  ducknng_request_raw(
    'wss://127.0.0.1:45456/_ducknng',
    from_hex('01000000000000000000000000000000000000000000'),
    1000,
    getvariable('tls_wss_id')::UBIGINT
  )
);

SELECT ducknng_stop_server('wss_demo');
SELECT ducknng_drop_tls_config(getvariable('tls_wss_id')::UBIGINT);
```

### REQ/REP `EXEC` via `nanonext` as an interop example

This example is easier to follow as a short sequence: define the frame helpers, define the retry helpers, start a local server, run a few remote `exec` requests, inspect the same table locally, and then clean up.

#### Frame encoding and decoding helpers

```{r exec-example-frame-helpers}
# Little-endian helpers for the versioned ducknng RPC envelope.
write_u32le <- function(x) {
  writeBin(as.integer(x), raw(), size = 4L, endian = "little")
}

write_u64le <- function(x) {
  x <- as.double(x)
  c(write_u32le(x %% 2^32), write_u32le(floor(x / 2^32)))
}

read_u32le_bytes <- function(x) {
  sum(as.double(as.integer(x)) * 256^(0:3))
}

read_u64le_bytes <- function(x) {
  read_u32le_bytes(x[1:4]) + 2^32 * read_u32le_bytes(x[5:8])
}

read_u32le <- function(buf, offset) {
  read_u32le_bytes(buf[offset + 0:3])
}

read_u64le <- function(buf, offset) {
  read_u64le_bytes(buf[offset + 0:7])
}

# Encode the Arrow IPC payload carried by the built-in exec RPC method.
encode_ducknng_exec_payload <- function(sql, want_result = FALSE) {
  raw_con <- rawConnection(raw(), open = "r+")
  on.exit(close(raw_con), add = TRUE)

  nanoarrow::write_nanoarrow(
    data.frame(sql = sql, want_result = want_result),
    raw_con
  )

  rawConnectionValue(raw_con)
}

# Build a version 1 RPC call frame for the named method.
encode_ducknng_call_frame <- function(method, payload, flags = 0L) {
  method_name <- charToRaw(method)

  c(
    as.raw(1L),
    as.raw(1L),
    write_u32le(flags),
    write_u32le(length(method_name)),
    write_u32le(0),
    write_u64le(length(payload)),
    method_name,
    payload
  )
}

encode_ducknng_exec_request <- function(sql, want_result = FALSE) {
  encode_ducknng_call_frame(
    method = "exec",
    payload = encode_ducknng_exec_payload(sql, want_result = want_result)
  )
}

# Decode a reply envelope and parse its Arrow IPC payload when present.
decode_ducknng_exec_reply <- function(buf) {
  header_size <- 23L

  if (!is.raw(buf) || length(buf) < (header_size - 1L)) {
    stop("ducknng reply frame was empty or truncated", call. = FALSE)
  }

  name_len <- read_u32le(buf, 7)
  error_len <- read_u32le(buf, 11)
  payload_len <- read_u64le(buf, 15)

  name_start <- header_size
  error_start <- name_start + name_len
  payload_start <- error_start + error_len
  payload_end <- payload_start + payload_len - 1L

  payload <- if (payload_len > 0) buf[payload_start:payload_end] else raw()

  list(
    version = as.integer(buf[1]),
    type = as.integer(buf[2]),
    flags = read_u32le(buf, 3),
    method = rawToChar(buf[name_start:(error_start - 1L)]),
    error = if (error_len > 0) rawToChar(buf[error_start:(payload_start - 1L)]) else "",
    payload = payload,
    data = if (payload_len > 0) as.data.frame(nanoarrow::read_nanoarrow(payload)) else NULL
  )
}
```

#### REQ/REP retry helpers

```{r exec-example-retry-helpers}
# Send and receive with the same retry style used in mangoro examples.
send_ducknng_frame <- function(socket, frame, timeout_ms = 1000L, max_attempts = 20L) {
  attempt <- 1L
  send_result <- nanonext::send(socket, frame, mode = "raw", block = timeout_ms)

  while (nanonext::is_error_value(send_result) && attempt < max_attempts) {
    Sys.sleep(0.25)
    send_result <- nanonext::send(socket, frame, mode = "raw", block = timeout_ms)
    attempt <- attempt + 1L
  }

  send_result
}

recv_ducknng_frame <- function(socket, timeout_ms = 1000L, max_attempts = 20L) {
  attempt <- 1L
  response <- nanonext::recv(socket, mode = "raw", block = timeout_ms)

  while (
    (
      nanonext::is_error_value(response) ||
      !is.raw(response) ||
      length(response) < 22L
    ) &&
    attempt < max_attempts
  ) {
    Sys.sleep(0.25)
    response <- nanonext::recv(socket, mode = "raw", block = timeout_ms)
    attempt <- attempt + 1L
  }

  response
}

run_ducknng_req <- function(socket, frame, timeout_ms = 1000L) {
  send_ducknng_frame(socket, frame, timeout_ms = timeout_ms)
  decode_ducknng_exec_reply(
    recv_ducknng_frame(socket, timeout_ms = timeout_ms)
  )
}

# Wait for an IPC listener path to appear before dialing it.
wait_for_ducknng_listener <- function(path, timeout_secs = 10, interval_secs = 0.1) {
  deadline <- Sys.time() + timeout_secs

  while (Sys.time() < deadline) {
    if (file.exists(path)) {
      return(invisible(TRUE))
    }
    Sys.sleep(interval_secs)
  }

  stop("ducknng IPC listener did not become ready in time", call. = FALSE)
}
```

#### Start a local ducknng server and dial a req socket

```{r exec-example-setup}
# Start a ducknng server in this R session, then talk to it over a req socket.
ext_path <- normalizePath("build/release/ducknng.duckdb_extension")
ipc_path <- "/tmp/ducknng_readme_exec.ipc"
unlink(ipc_path)
ipc_url <- paste0("ipc://", ipc_path)

db_driver <- duckdb::duckdb(
  dbdir = ":memory:",
  config = list(allow_unsigned_extensions = "true")
)
db_con <- DBI::dbConnect(db_driver)

DBI::dbGetQuery(db_con, "SELECT 42 AS ok")
DBI::dbExecute(db_con, sprintf("LOAD '%s'", ext_path))
DBI::dbGetQuery(
  db_con,
  sprintf(
    "SELECT ducknng_start_server('sql_exec', '%s', 1, 134217728, 300000, 0::UBIGINT)",
    ipc_url
  )
)
DBI::dbGetQuery(db_con, "SELECT ducknng_register_exec_method()")

wait_for_ducknng_listener(ipc_path)
req <- nanonext::socket("req", dial = ipc_url)
```

#### Create and insert rows remotely

```{r exec-example-mutate}
# Create a table remotely and inspect the metadata reply.
create_reply <- run_ducknng_req(
  req,
  encode_ducknng_exec_request("CREATE TABLE ducknng_exec_demo(i INTEGER)")
)
create_reply$data

# Insert rows remotely and inspect the second metadata reply.
insert_reply <- run_ducknng_req(
  req,
  encode_ducknng_exec_request("INSERT INTO ducknng_exec_demo VALUES (42), (99)")
)
insert_reply$data
```

#### Query rows remotely with `want_result = TRUE`

```{r exec-example-query}
select_reply <- run_ducknng_req(
  req,
  encode_ducknng_exec_request(
    "SELECT i, i > 50 AS gt_50 FROM ducknng_exec_demo ORDER BY i",
    want_result = TRUE
  )
)
select_reply$data
```

#### Inspect the same table locally through DuckDB

```{r exec-example-local}
server_rows <- DBI::dbGetQuery(db_con, "SELECT * FROM ducknng_exec_demo ORDER BY i")
server_rows
```

#### Stop the server and clean up

```{r exec-example-cleanup}
DBI::dbGetQuery(db_con, "SELECT ducknng_stop_server('sql_exec')")
close(req)
DBI::dbDisconnect(db_con)
unlink(ipc_path)
```

## Deployment and admission policy

`ducknng` defaults to low-level, default-open transport behavior. The application or deployment decides where the trust boundary is. For anything beyond process-local, filesystem-local, or loopback-only development, set the boundary deliberately before exposing a service.

A strong direct-exposure baseline is:

1. use `https://`, `wss://`, or `tls+tcp://` rather than plaintext network carriers;
2. create the TLS handle with `auth_mode = 2` so peers must present a verified client certificate;
3. configure a private CA or otherwise narrow certificate issuance so the trust root is not broader than the application;
4. set exact peer-identity and/or IP/CIDR allowlists before exposure with `ducknng_set_tls_peer_allowlist(...)` and the optional `ip_allowlist_json` startup argument;
5. set basic service resource limits such as `ducknng_set_service_limits(name, max_open_sessions[, max_active_pipes[, max_inflight_requests[, max_sessions_per_peer_identity]]])` before exposing query sessions;
6. install a service-level SQL authorizer with `ducknng_set_service_authorizer(...)` when admission depends on deployment tables, tenants, method names, or HTTP metadata;
7. optionally update a running service dynamically with `ducknng_set_service_peer_allowlist(...)`, `ducknng_set_service_ip_allowlist(...)`, `ducknng_set_service_limits(...)`, and `ducknng_set_service_authorizer(...)`.

Free-form SQL execution is intentionally a deployment-owned capability. `ducknng` keeps its own generated SQL injection-safe, but it does not make arbitrary user SQL safe: `exec` and query sessions run DuckDB SQL as the service process on the configured DuckDB connection. Do not expose `exec` or unrestricted query sessions to callers who are not already authorized to run that SQL. Use these profiles as a starting point:

- **Local development / single-user IPC**: `inproc://`, loopback, or `ipc://` under a private directory; `exec` may be enabled for convenience inside that local boundary.
- **Trusted service mesh**: `tls+tcp://`, `wss://`, or `https://` with mTLS, exact peer allowlists, IP/CIDR allowlists where useful, and service limits; register `exec` with `ducknng_register_exec_method(true)` or leave it disabled.
- **Shared or semi-trusted clients**: prefer query sessions plus a short SQL authorizer; use deployment-level DuckDB/process controls for filesystem, extension loading, outbound network, and attachments.
- **Public/untrusted internet**: do not expose raw DuckDB SQL directly. Put `ducknng` behind a gateway that authenticates, rate-limits, and maps users to fixed application operations, or build a separate route layer that does not forward arbitrary SQL text.

When writing SQL authorizers, read request fields from `ducknng_auth_context()` and avoid string-concatenating untrusted HTTP headers, method names, or payload text into new SQL. Keep the callback short and side-effect-light.

The relevant identity strings currently come from verified TLS peer metadata and are SAN-first, CN-fallback: `tls:san:<value>` or `tls:cn:<common-name>`. Peer identity allowlists are exact-match. IP allowlists are parsed once into IPv4/IPv6 literal or CIDR rules such as `127.0.0.1/32`, `10.0.0.0/8`, or `::1/128`. `NULL` or an empty SQL string clears an allowlist and returns to default-open admission subject to TLS/method/session policy. An empty JSON array, `[]`, is active deny-all.

Example shape:

```sql
-- Set a default allowlist on the TLS handle before starting services with it.
SELECT ducknng_set_tls_peer_allowlist(
  1::UBIGINT,
  '["tls:san:client-a.example","tls:san:client-b.example"]'
);

-- Start with an IP/CIDR allowlist already installed before the listener accepts traffic.
SELECT ducknng_start_server(
  'sql_http',
  'http://127.0.0.1:18444/_ducknng',
  1,
  134217728,
  300000,
  0::UBIGINT,
  '["127.0.0.1/32"]'
);

-- Later, rotate or tighten admission for a running service.
SELECT ducknng_set_service_peer_allowlist(
  'sql_rpc',
  '["tls:san:client-b.example"]'
);
SELECT ducknng_set_service_ip_allowlist(
  'sql_http',
  '["127.0.0.1/32","10.0.0.0/8"]'
);

-- Set a basic service-side resource limit for query sessions.
SELECT ducknng_set_service_limits('sql_http', 16::UBIGINT);

-- Add a flexible SQL callback when policy depends on request context.
SELECT ducknng_set_service_authorizer(
  'sql_http',
  'SELECT remote_ip = ''127.0.0.1'' AND rpc_method = ''manifest'' AS allow,
          403 AS status,
          ''not authorized'' AS reason
   FROM ducknng_auth_context()'
);
```

Transport-specific deployment notes:

| Surface | Accepted schemes | TLS handle accepted on |
| --- | --- | --- |
| `ducknng_start_server(...)` | `inproc://`, `ipc://`, `tcp://`, `tls+tcp://`, `ws://`, `wss://`, `http://`, `https://` | `tls+tcp://`, `wss://`, `https://` |
| synchronous RPC/session helpers | NNG schemes plus `http://`, `https://` | `tls+tcp://`, `wss://`, `https://` |
| raw RPC AIO helpers | NNG schemes plus `http://`, `https://` | `tls+tcp://`, `wss://`, `https://` |
| generic socket APIs | `inproc://`, `ipc://`, `tcp://`, `tls+tcp://`, `ws://`, `wss://` | `tls+tcp://`, `wss://` |
| HTTP client helpers (`ducknng_ncurl(...)`, `ducknng_ncurl_aio(...)`, `ducknng_ncurl_table(...)`) | `http://`, `https://` | `https://` |

Supplying a non-zero TLS handle on a non-TLS URL is rejected rather than silently ignored. HTTP/HTTPS URLs are rejected by the generic NNG socket layer, and NNG URLs are rejected by HTTP-only helpers.

- `inproc://` is a same-process boundary. It is useful for embedded testing/composition, but it is not network authentication.
- `ipc://` relies on filesystem or named-pipe permissions. Avoid world-writable shared paths for multi-user deployments.
- plaintext `tcp://`, `ws://`, and `http://` do not provide cryptographic peer identity. Use them only on loopback, a private network you already trust, or behind a reverse proxy/firewall/VPN that performs admission itself.
- `http://` is still useful as a local or reverse-proxy upstream carrier. For public HTTP-style deployment, expose `https://` directly with mTLS or place `ducknng` behind a proxy such as Caddy, nginx, Envoy, or a cloud load balancer that terminates TLS and enforces authentication/rate limits before forwarding to loopback `http://` or `ipc://`.
- `tls+tcp://` and `wss://` are NNG protocol transports. They use NNG socket protocols and NNG pipes, not the HTTP carrier semantics.
- `https://` is the HTTP frame carrier over TLS. It shares the same framed RPC methods and admission policy, but HTTP status codes such as `403` can appear before RPC dispatch.

IP allowlisting is a useful lighter-weight gate when client certificates are too much operational overhead, but it is not the same as cryptographic client identity. HTTPS itself does not standardize an application IP allowlist; IP admission is normally implemented by the listener, firewall, reverse proxy, load balancer, or service mesh. `ducknng` now includes its own service-level IP/CIDR allowlist for direct listeners. For NNG protocol transports, it uses `NNG_OPT_REMADDR` from the NNG pipe and checks parsed binary CIDR rules during `ADD_PRE` and again at dispatch. For HTTP/HTTPS, it reads the connection remote address from the NNG supplemental HTTP connection and returns HTTP `403` before RPC dispatch when the address is not admitted. A front door such as Caddy, nginx, Envoy, cloud firewall/security groups, Kubernetes network policy, or a similar proxy remains useful for public deployments because it can combine IP/CIDR policy with rate limits, request logging, WAF rules, and ordinary web authentication before forwarding to loopback `http://`, `https://`, `ipc://`, or an internal NNG service.

For NNG transports, admission uses NNG's own pipe notification primitive: `nng_pipe_notify(..., NNG_PIPE_EV_ADD_PRE, ...)` lets `ducknng` inspect verified TLS pipe metadata and remote-address metadata and close a non-admitted pipe before it is added to the socket. For HTTP/HTTPS, NNG's supplemental HTTP API does not expose the same public protocol-socket pipe admission hook, so `ducknng` checks the same policy before RPC dispatch and returns HTTP `403` for non-admitted peers or remote addresses.

The same pipe-event family is useful beyond admission. NNG exposes `ADD_PRE`, `ADD_POST`, and `REM_POST` pipe events; nanonext surfaces related tools as `pipe_notify()`, `monitor()`, and `read_monitor()`. `ducknng` now records a bounded per-service NNG pipe event stream and exposes it with `ducknng_read_monitor(name, after_seq, max_events)`. It exposes ring/counter metadata with `ducknng_monitor_status(name)` and current active NNG pipes with `ducknng_list_pipes(name)`. The event stream includes event sequence, timestamp, service, transport, pipe id, admission result for `ADD_PRE`, denial reason when the fast C path has one, remote address, and verified peer identity when present; the active-pipe table gives the current pipe set for routing and membership decisions. These primitives are a natural foundation for telemetry, connection counts, connection churn, per-service pipe event streams, presence/worker membership, dynamic routing, backpressure-aware scheduling, mesh-style DuckDB service discovery, routing/forwarding demos, and other application-level event streams. HTTP/HTTPS framed RPC currently uses the same admission policy but does not expose NNG protocol-socket pipe events; a broader HTTP server framework can reuse the same request context and monitor concepts without changing the framed RPC wire protocol.

## Status

What is sealed and runnable in 0.1.0: NNG transports and socket patterns; first-class one-shot AIO across socket send/recv, ncurl HTTP, and unary RPC; framed RPC (`manifest`, opt-in `exec`, raw unary, query sessions); low-level HTTP routes with exact, prefix, or template matching plus SQL response rows, request-context/body helpers, named request accessors, and one-row response builders; fast C admission (mTLS, exact peer-identity allowlists, IP/CIDR allowlists, service limits); SQL authorizer callbacks; bounded per-service pipe-event monitor and active-pipe snapshot; body codec layer with built-in providers and user-registered codec hooks. See `function_catalog/functions.md` for the exact surface and `NEWS.md` for landed changes.

Intentionally deferred:

- Route-local auth policies, static asset serving, streaming HTTP route helpers, worker lifecycle management, and richer web-toolkit conveniences beyond the current low-level route framework.
- CSV/TSV/Parquet body parsers beyond the safe BLOB fallback.
- Full SQL-side decoding of session `fetch` Arrow batch BLOBs into a table-function path.
- Routing/forwarding and mesh scheduling as sealed APIs — the pipe monitor and active-pipe snapshot are the membership/telemetry foundation that comes first.

## References

- [NNG](https://nng.nanomsg.org/) for the underlying messaging library and transport family.
- [`nanonext`](https://github.com/r-lib/nanonext) for the main client/server ergonomics reference.
- [`mangoro`](https://github.com/sounkou-bioinfo/mangoro) for the thin-envelope + Arrow IPC RPC direction.
- [DuckDB C API](https://duckdb.org/docs/stable/clients/c/api) for the extension and SQL integration boundary.
- [Apache Arrow IPC](https://arrow.apache.org/docs/format/Columnar.html#serialization-and-interprocess-communication-ipc) for the tabular payload format.
