---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::knit_engines$set(sql = function(options) {
  code <- paste(options$code, collapse = "\n")
  extension_path <- "build/release/ducknng.duckdb_extension"
  extension_path <- gsub("'", "''", extension_path, fixed = TRUE)
  if (!identical(options$autoload, FALSE)) {
    code <- paste(sprintf("LOAD '%s';", extension_path), code, sep = "\n")
  }
  if (isTRUE(options$eval)) {
    run_code <- paste(".mode table", code, sep = "\n")
    out <- system2("duckdb", "-unsigned", input = run_code, stdout = TRUE, stderr = TRUE)
    status <- attr(out, "status")
    if (!is.null(status) && status != 0) stop(paste(out, collapse = "\n"))
    knitr::engine_output(options, code, out)
  } else {
    knitr::engine_output(options, code, NULL)
  }
})
```

# ducknng

`ducknng` is a pure C DuckDB extension that turns a DuckDB process into an [NNG](https://nng.nanomsg.org/) endpoint — server, client, or both — and adds a small framed RPC layer so a SQL session can call another DuckDB session across `inproc://`, `ipc://`, `tcp://`, `tls+tcp://`, `ws://`, `wss://`, `http://`, or `https://`. RPC row payloads ride on Arrow IPC.

It draws on two R packages. [`nanonext`](https://github.com/r-lib/nanonext) supplies the socket and AIO ergonomics; [`mangoro`](https://github.com/sounkou-bioinfo/mangoro) supplies the thin versioned envelope. The DuckDB integration keeps generated SQL injection-safe but assumes the deployment owns the trust boundary — see **Deployment and admission policy** below.

### How the pieces stack

The extension is one runtime with four layers; later sections walk through each in turn.

1. **Transport.** NNG sockets and listeners across the full transport family. Synchronous send/recv plus first-class one-shot AIO handles. Per-service pipe-event telemetry (`ducknng_read_monitor(...)`, `ducknng_list_pipes(...)`).
2. **Framed RPC.** A versioned envelope carries Arrow IPC table payloads or JSON control text. Servers come up with `ducknng_start_server(...)`, and the listen URL scheme chooses NNG or HTTP/HTTPS. HTTP/HTTPS stays on the same server entrypoint with `contexts = 1`. Clients call by URL; synchronous helpers route by scheme, AIO helpers expose the same calls as futures.
3. **Policy.** Manifest-driven method registry with opt-in `exec`. Fast C admission for required mTLS, exact peer-identity allowlists, IP/CIDR allowlists, and service limits. Optional SQL authorizers run at the request boundary and read context through `ducknng_auth_context()`. Sessions are bearer-token owned, additionally bound to verified mTLS identity when present.
4. **Codecs.** `ducknng_parse_body(...)` and `ducknng_ncurl_table(...)` parse content-type-tagged BLOBs. Built-in providers cover JSON, Arrow IPC, ducknng frames, and a text/raw fallback. Deployments can override or extend the set with `ducknng_register_codec(content_type, function_name)`; the function takes `BLOB` and returns `VARCHAR`.

`ducknng` is intentionally low-level. Long-lived runtime handles — servers, sockets, AIO, TLS configs, query sessions — are **manually managed**. Stop, close, or drop them explicitly; runtime teardown is fallback cleanup, not the primary lifecycle.

## Quick tour

Build the extension:

```sh
make configure
make release
```

Load it, start an IPC service, ask the server its name through the built-in `manifest` method, and stop it again:

```{sql eval=TRUE, autoload=FALSE, collapse=FALSE, comment=""}
LOAD 'build/release/ducknng.duckdb_extension';

-- args: name, listen URL, REP-context count, recv_max_bytes,
--       session_idle_ms, tls_config_id (0 = plaintext).
SELECT ducknng_start_server(
  'hello', 'ipc:///tmp/ducknng_hello.ipc',
  1, 134217728, 300000, 0::UBIGINT
);

-- The manifest column is JSON text held as VARCHAR; cast to JSON to drill in.
SELECT json_extract_string(manifest::JSON, '$.server.name') AS server_name,
       json_array_length(json_extract(manifest::JSON, '$.methods')) AS methods
FROM ducknng_get_rpc_manifest('ipc:///tmp/ducknng_hello.ipc', 0::UBIGINT)
WHERE ok;

SELECT ducknng_stop_server('hello');
```

Point `ducknng_start_server(...)` at an `http://` or `https://` URL with `contexts = 1` and the same RPC methods are mounted on the HTTP carrier. The **Examples** section walks through transport patterns, AIO, the client surface, body codecs, and sessions in that order.

## Lifetime and manual cleanup

DuckDB's extension API does give `ducknng` destroy callbacks for internal function-registration, bind, and init state. What DuckDB SQL does **not** give this project is a general-purpose GC/finalizer model for long-lived SQL handles like R external pointer finalizers. So at the public SQL surface, long-lived handles are manually managed.

You should explicitly clean up what you create:

- stop servers with `ducknng_stop_server(...)`
- close sockets with `ducknng_close_socket(...)`
- drop aio handles with `ducknng_aio_drop(...)`
- drop TLS config handles with `ducknng_drop_tls_config(...)`
- close query sessions with `ducknng_close_query(...)`

Important details:

- `ducknng_aio_cancel(...)` is cancellation control, not the destructor
- `ducknng_cancel_query(...)` is best-effort session control, not a replacement for the normal explicit close path
- runtime teardown is fallback cleanup, not the primary lifecycle model for a long-lived DuckDB process

For the deeper write-ups see `docs/lifetime.md`, `docs/protocol.md`, `docs/manifest.md`, `docs/security.md`, `docs/registry.md`, `docs/transports.md`, `docs/http.md`, `docs/codecs.md`, and `docs/types.md`. The non-sealed routing demo lives in `docs/mesh_routing_demo.md`; `NEWS.md` summarizes landed changes. TLS supports both file-backed and in-memory PEM material (`ducknng_tls_config_from_files(...)`, `ducknng_tls_config_from_pem(...)`, `ducknng_self_signed_tls_config(...)`); `auth_mode = 2` enables required mTLS peer verification for `tls+tcp://`, `wss://`, and `https://` services.

## Function catalog

<details>
<summary>Expand the full generated function catalog</summary>

```{r function-catalog, echo=FALSE, results='asis'}
knitr::asis_output(paste(readLines("function_catalog/functions.md"), collapse = "\n"))
```

</details>

## Examples

### Start an IPC listener and inspect the registry

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT ducknng_start_server(
  'sql0',                         -- service name
  'ipc:///tmp/ducknng_sql0.ipc', -- listen URL
  1,                              -- REP contexts
  134217728,                      -- recv_max_bytes
  300000,                         -- session_idle_ms
  0                               -- tls_config_id (0 means plaintext)
);

SELECT name, listen, contexts, running, sessions
FROM ducknng_list_servers();

SELECT ducknng_stop_server('sql0');
```

### Request multiple REP contexts on one REP socket

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT ducknng_start_server(
  'sql_multi',                    -- service name
  'ipc:///tmp/ducknng_sql_multi.ipc', -- listen URL
  3,                              -- REP contexts
  134217728,                      -- recv_max_bytes
  300000,                         -- session_idle_ms
  0                               -- tls_config_id (0 means plaintext)
);

SELECT name, contexts, running
FROM ducknng_list_servers()
WHERE name = 'sql_multi';

SELECT ducknng_stop_server('sql_multi');
```

### DuckDB can also act as a client

```{sql eval=TRUE, collapse=FALSE, comment=""}
-- Start a local ducknng service that the following client examples will talk to.
SELECT ducknng_start_server(
  'sql_client_demo',              -- service name
  'ipc:///tmp/ducknng_sql_client_demo.ipc', -- listen URL
  1,                              -- REP contexts
  134217728,                      -- recv_max_bytes
  300000,                         -- session_idle_ms
  0                               -- tls_config_id (0 means plaintext)
);

-- The default RPC surface is minimal: manifest is built in, exec is opt-in.
WITH manifest_row AS (
  SELECT manifest
  FROM ducknng_get_rpc_manifest('ipc:///tmp/ducknng_sql_client_demo.ipc', 0::UBIGINT)
  WHERE ok
)
SELECT json_extract_string(manifest::JSON, '$.server.name') AS server_name,
       json_array_length(json_extract(manifest::JSON, '$.methods')) AS method_count,
       position('"name":"exec"' IN manifest) > 0 AS has_exec
FROM manifest_row;
SELECT name, family, response_mode, requires_auth, disabled
FROM ducknng_list_methods()
ORDER BY name;

-- Register exec explicitly before exposing SQL execution over RPC.
-- Pass TRUE to require verified transport-derived peer identity for exec dispatch.
SELECT ducknng_register_exec_method(false) AS registered_exec;
SELECT name, family, response_mode, requires_auth, disabled
FROM ducknng_list_methods()
ORDER BY name;
WITH manifest_row AS (
  SELECT manifest
  FROM ducknng_get_rpc_manifest('ipc:///tmp/ducknng_sql_client_demo.ipc', 0::UBIGINT)
  WHERE ok
)
SELECT json_extract_string(manifest::JSON, '$.server.name') AS server_name,
       json_array_length(json_extract(manifest::JSON, '$.methods')) AS method_count,
       position('"name":"exec"' IN manifest) > 0 AS has_exec
FROM manifest_row;

-- RPC helper: run non-row statements and keep errors in-band.
SELECT * FROM ducknng_run_rpc('ipc:///tmp/ducknng_sql_client_demo.ipc', 'CREATE TABLE IF NOT EXISTS client_side_demo(i INTEGER)', 0::UBIGINT);
SELECT * FROM ducknng_run_rpc('ipc:///tmp/ducknng_sql_client_demo.ipc', 'INSERT INTO client_side_demo VALUES (10), (11)', 0::UBIGINT);

-- RPC helper: fetch row results through the unary query path.
SELECT *
FROM ducknng_query_rpc(
  'ipc:///tmp/ducknng_sql_client_demo.ipc',          -- url
  'SELECT i, i > 10 AS gt_10 FROM client_side_demo ORDER BY i', -- sql
  0::UBIGINT                                        -- tls_config_id
);

-- The Arrow IPC row path also carries temporal, decimal, list, and struct values.
SELECT d = DATE '2024-01-02' AS date_ok,
       ts = TIMESTAMP '2024-01-02 03:04:05.123456' AS ts_ok,
       dec = 123.45::DECIMAL(10,2) AS decimal_ok,
       xs[2] IS NULL AND xs[3] = 3 AS list_ok,
       st.a = 7 AND st.b = 'bee' AS struct_ok
FROM ducknng_query_rpc(
  'ipc:///tmp/ducknng_sql_client_demo.ipc',          -- url
  'SELECT DATE ''2024-01-02'' AS d,
          TIMESTAMP ''2024-01-02 03:04:05.123456'' AS ts,
          123.45::DECIMAL(10,2) AS dec,
          [1, NULL, 3]::INTEGER[] AS xs,
          {''a'': 7::INTEGER, ''b'': ''bee''} AS st', -- sql
  0::UBIGINT                                        -- tls_config_id
);

-- Body codec helpers parse content-type-tagged BLOBs into SQL tables.
SELECT provider, output
FROM ducknng_list_codecs()
ORDER BY provider;

SELECT a, b
FROM ducknng_parse_body(
  '[{"a":1,"b":"x"},{"a":2,"b":"y"}]'::BLOB,
  'application/json; charset=utf-8'
)
ORDER BY a;

-- Primitive transport layer: open a req socket handle, dial it, and inspect the registry.
SELECT ducknng_open_socket('req');
SELECT ducknng_dial_socket(1, 'ipc:///tmp/ducknng_sql_client_demo.ipc', 1000, 0::UBIGINT);
SELECT * FROM ducknng_list_sockets();

-- Primitive transport layer: send the built-in manifest request frame.
-- The hex literal here is the current wire-format request for the always-on manifest method.
SELECT ok, error, octet_length(payload) > 0 AS has_payload
FROM ducknng_request_socket(
  1::UBIGINT,                                        -- socket_id
  from_hex('01000000000000000000000000000000000000000000'), -- manifest request frame
  1000                                               -- timeout_ms
);

SELECT ok, error, octet_length(payload) > 0 AS has_payload
FROM ducknng_request(
  'ipc:///tmp/ducknng_sql_client_demo.ipc',          -- url
  from_hex('01000000000000000000000000000000000000000000'), -- manifest request frame
  1000,                                              -- timeout_ms
  0::UBIGINT                                         -- tls_config_id
);

-- Raw scalar forms now mean raw reply frames as BLOBs.
SELECT substr(
  hex(
    ducknng_request_socket_raw(
      1,                                             -- socket_id
      from_hex('01000000000000000000000000000000000000000000'), -- manifest request frame
      1000                                           -- timeout_ms
    )
  ),
  1,
  28
);

-- Decode raw manifest and exec replies into envelope fields plus extracted text payload.
SELECT ok, version, type_name, name, position('"name":"exec"' IN payload_text) > 0
FROM ducknng_decode_frame(ducknng_get_rpc_manifest_raw('ipc:///tmp/ducknng_sql_client_demo.ipc', 0::UBIGINT));

SELECT ok, type_name, name
FROM ducknng_decode_frame(
  ducknng_run_rpc_raw('ipc:///tmp/ducknng_sql_client_demo.ipc', 'CREATE TABLE IF NOT EXISTS client_side_demo(i INTEGER)', 0::UBIGINT)
);

-- The generic raw request helper can be decoded the same way.
SELECT ok, version, type_name, name, position('"name":"exec"' IN payload_text) > 0 AS has_exec
FROM ducknng_decode_frame(
  ducknng_request_raw('ipc:///tmp/ducknng_sql_client_demo.ipc', from_hex('01000000000000000000000000000000000000000000'), 1000, 0::UBIGINT)
);

-- Close the client socket handle and stop the demo server.
SELECT ducknng_close_socket(1);
SELECT ducknng_stop_server('sql_client_demo');
```

### Use `ducknng_ncurl()` against a local `nanonext` HTTPS server

A tiny local R `nanonext` HTTPS server is started during README rendering so this example runs without depending on the public internet. The core server definition is shown below because the transport helper and its carrier behavior are part of the example rather than hidden setup. The hidden setup only launches the script in the background and manages the temporary PID, log, and CA files needed while rendering. `ducknng_ncurl(...)` remains the low-level HTTP/HTTPS transport primitive even though the higher-level synchronous request, RPC, and session helpers now auto-route over `http://` and `https://` when the URL scheme selects that carrier.

```{r http-client-demo-setup, echo=FALSE, results='asis'}
library(nanonext)
pidfile <- "/tmp/ducknng_readme_http_demo.pid"
script <- "/tmp/ducknng_readme_http_demo.R"
logfile <- "/tmp/ducknng_readme_http_demo.log"
cafile <- "/tmp/ducknng_readme_http_demo_ca.pem"
if (file.exists(pidfile)) {
  pid <- suppressWarnings(as.integer(readLines(pidfile, warn = FALSE)))
  if (!is.na(pid)) tools::pskill(pid)
  unlink(pidfile)
}
http_demo_code <- paste(
  "library(nanonext)",
  "",
  "cert <- write_cert(cn = '127.0.0.1')",
  sprintf("writeLines(cert$client[[1]], %s)", deparse(cafile)),
  "",
  "server <- http_server(",
  "  url = 'https://127.0.0.1:18443',",
  "  handlers = list(",
  "    handler('/hello', function(req) {",
  "      list(",
  "        status = 200L,",
  "        headers = c(",
  "          'Content-Type' = 'text/plain',",
  "          'X-Test' = 'hello'",
  "        ),",
  "        body = 'hello from nanonext https server'",
  "      )",
  "    }),",
  "    handler('/echo', function(req) {",
  "      list(",
  "        status = 200L,",
  "        headers = c(",
  "          'Content-Type' = (",
  "            req$headers[['Content-Type']] %||% 'application/octet-stream'",
  "          ),",
  "          'X-Test' = 'echo'",
  "        ),",
  "        body = req$body",
  "      )",
  "    }, method = 'POST')",
  "  ),",
  "  tls = tls_config(server = cert$server)",
  ")",
  sep = "\n"
)
cat("```r\n", http_demo_code, "\n```\n\n", sep = "")
launch_code <- paste(
  http_demo_code,
  "",
  "library(later)",
  "attr(server, 'start')()",
  "repeat later::run_now(Inf)",
  sep = "\n"
)
writeLines(launch_code, script)
status <- system(sprintf(
  "Rscript %s > %s 2>&1 & echo $! > %s",
  shQuote(script),
  shQuote(logfile),
  shQuote(pidfile)
))
if (status != 0 || !file.exists(pidfile)) {
  stop("failed to start the README HTTP demo server")
}
pid <- suppressWarnings(as.integer(readLines(pidfile, warn = FALSE)))
if (is.na(pid) || system(sprintf("kill -0 %d >/dev/null 2>&1", pid)) != 0) {
  log_text <- if (file.exists(logfile)) paste(readLines(logfile, warn = FALSE), collapse = "\n") else ""
  stop("README HTTP demo server did not stay running\n", log_text)
}
ready <- FALSE
for (attempt in seq_len(20)) {
  client_tls <- tryCatch(
    if (file.exists(cafile)) tls_config(client = cafile) else NULL,
    error = function(e) NULL
  )
  probe <- tryCatch(
    if (!is.null(client_tls)) ncurl("https://127.0.0.1:18443/hello", timeout = 500L, tls = client_tls) else NULL,
    error = function(e) NULL
  )
  if (!is.null(probe) && identical(probe$status, 200L)) {
    ready <- TRUE
    break
  }
  Sys.sleep(0.25)
}
if (!ready) {
  log_text <- if (file.exists(logfile)) paste(readLines(logfile, warn = FALSE), collapse = "\n") else ""
  stop("README HTTP demo server did not become ready\n", log_text)
}
```

```{sql eval=TRUE, collapse=FALSE, comment=""}
-- Register a client TLS handle that trusts the self-signed HTTPS demo server.
SELECT ducknng_tls_config_from_files(
  NULL,                                     -- cert_key_file
  '/tmp/ducknng_readme_http_demo_ca.pem',   -- ca_file
  NULL,                                     -- password
  2                                         -- auth_mode = require certificate validation
);

-- Call /hello through ducknng_ncurl(): NULL method means GET and NULL body means no request body.
SELECT ok, status, error, body_text
FROM ducknng_ncurl(
  'https://127.0.0.1:18443/hello', -- url
  NULL,                            -- method
  NULL,                            -- headers_json
  NULL,                            -- body
  2000,                            -- timeout_ms
  1::UBIGINT                       -- tls_config_id
);

-- POST can send raw bytes while still exposing HTTP headers and status in-band.
SET VARIABLE echo_headers =
  '[{"name":"Content-Type","value":"application/octet-stream"}]';

SELECT
  ok,
  status,
  error,
  hex(body) AS body_hex,
  position('X-Test' IN headers_json) > 0 AS has_header
FROM ducknng_ncurl(
  'https://127.0.0.1:18443/echo', -- url
  'POST',                         -- method
  getvariable('echo_headers'),    -- headers_json
  from_hex('01020304'),           -- body
  2000,                           -- timeout_ms
  1::UBIGINT                      -- tls_config_id
);

-- Drop the temporary client TLS handle after the HTTPS checks complete.
SELECT ducknng_drop_tls_config(1);
```

```{r http-client-demo-cleanup, include=FALSE}
pidfile <- "/tmp/ducknng_readme_http_demo.pid"
if (file.exists(pidfile)) {
  pid <- suppressWarnings(as.integer(readLines(pidfile, warn = FALSE)))
  if (!is.na(pid)) tools::pskill(pid)
  unlink(pidfile)
}
unlink(c(
  "/tmp/ducknng_readme_http_demo.R",
  "/tmp/ducknng_readme_http_demo.log",
  "/tmp/ducknng_readme_http_demo_ca.pem"
))
```

### Start `ducknng` on `http://` and use the routed helpers directly

`ducknng_start_server(...)` mounts the existing framed RPC surface at the exact path encoded in the listen URL, and the higher-level synchronous request, RPC, and session helpers keep the same names they already use for NNG URLs while switching carriers automatically from the scheme. For `http://` and `https://`, use `contexts = 1`.

```{sql eval=TRUE, collapse=FALSE, comment=""}
SELECT ducknng_start_server(
  'http_demo',
  'http://127.0.0.1:18444/_ducknng',
  1,
  134217728,
  300000,
  0::UBIGINT
);

WITH manifest_row AS (
  SELECT manifest
  FROM ducknng_get_rpc_manifest('http://127.0.0.1:18444/_ducknng', 0::UBIGINT)
  WHERE ok
)
SELECT json_extract_string(manifest::JSON, '$.server.name') AS server_name,
       json_array_length(json_extract(manifest::JSON, '$.methods')) AS method_count,
       position('"name":"manifest"' IN manifest) > 0 AS has_manifest
FROM manifest_row;

SET VARIABLE http_demo_frame = (
  SELECT body
  FROM ducknng_ncurl(
    'http://127.0.0.1:18444/_ducknng',
    'POST',
    '[{"name":"Content-Type","value":"application/vnd.ducknng.frame"}]',
    from_hex('01000000000000000000000000000000000000000000'),
    2000,
    0::UBIGINT
  )
);

SELECT ok, type_name, name
FROM ducknng_decode_frame(getvariable('http_demo_frame'));

SELECT ducknng_stop_server('http_demo');
```

### Launch raw socket send/recv airos and inspect send status explicitly

```{sql eval=TRUE, collapse=FALSE, comment=""}
-- Open one listening pair socket and one dialed peer.
-- SET VARIABLE keeps setup handles out of the rendered output.
SET VARIABLE pair_a = ducknng_open_socket('pair');
SET VARIABLE pair_listen_ok = ducknng_listen_socket(
  getvariable('pair_a')::UBIGINT,
  'ipc:///tmp/ducknng_sql_pair_aio_demo.ipc',
  134217728,
  0::UBIGINT
);
SET VARIABLE pair_b = ducknng_open_socket('pair');
SET VARIABLE pair_dial_ok = ducknng_dial_socket(
  getvariable('pair_b')::UBIGINT,
  'ipc:///tmp/ducknng_sql_pair_aio_demo.ipc',
  1000,
  0::UBIGINT
);

-- Start one async receive and one async send.
CREATE TEMP TABLE pair_recv AS
SELECT ducknng_recv_socket_raw_aio(getvariable('pair_a')::UBIGINT, 1000) AS recv_aio;

CREATE TEMP TABLE pair_send AS
SELECT ducknng_send_socket_raw_aio(
  getvariable('pair_b')::UBIGINT,
  from_hex('6173796e632072657175657374'),
  1000
) AS send_aio;

-- Collect the terminal send result. Send-only airos succeed with frame = NULL.
SELECT aio_id, ok, frame IS NULL AS no_frame
FROM ducknng_aio_collect((SELECT list_value(send_aio) FROM pair_send), 1000);

-- Collect the received raw frame from the peer side.
SELECT aio_id, ok, hex(frame) = '6173796E632072657175657374' AS got_payload
FROM ducknng_aio_collect((SELECT list_value(recv_aio) FROM pair_recv), 1000);

-- Inspect whether the send phase completed successfully.
SELECT kind, state, phase, send_done, send_ok, recv_done, recv_ok IS NULL AS recv_ok_is_null
FROM ducknng_aio_status((SELECT send_aio FROM pair_send));

-- Clean up the terminal aio handles, temp state, and socket handles.
SET VARIABLE pair_drop_send = ducknng_aio_drop((SELECT send_aio FROM pair_send));
SET VARIABLE pair_drop_recv = ducknng_aio_drop((SELECT recv_aio FROM pair_recv));
DROP TABLE pair_send;
DROP TABLE pair_recv;
SET VARIABLE pair_close_b = ducknng_close_socket(getvariable('pair_b')::UBIGINT);
SET VARIABLE pair_close_a = ducknng_close_socket(getvariable('pair_a')::UBIGINT);
```

### Push one raw message through `push` / `pull`

```{sql eval=TRUE, collapse=FALSE, comment=""}
-- Open one pull socket, listen on it, then dial it from a push peer.
SET VARIABLE pull_socket = ducknng_open_socket('pull');
SET VARIABLE pull_listen_ok = ducknng_listen_socket(
  getvariable('pull_socket')::UBIGINT,
  'ipc:///tmp/ducknng_sql_pushpull_demo.ipc',
  134217728,
  0::UBIGINT
);
SET VARIABLE push_socket = ducknng_open_socket('push');
SET VARIABLE push_dial_ok = ducknng_dial_socket(
  getvariable('push_socket')::UBIGINT,
  'ipc:///tmp/ducknng_sql_pushpull_demo.ipc',
  1000,
  0::UBIGINT
);

-- Receive asynchronously on pull so the send side can return immediately.
CREATE TEMP TABLE pushpull_recv AS
SELECT ducknng_recv_socket_raw_aio(getvariable('pull_socket')::UBIGINT, 1000) AS recv_aio;

SET VARIABLE pushpull_sent = ducknng_send_socket_raw(
  getvariable('push_socket')::UBIGINT,
  from_hex('707573682d70756c6c'),
  1000
);

SELECT getvariable('pushpull_sent')::BOOLEAN AS sent,
       ok,
       hex(frame) = '707573682D70756C6C' AS got_payload
FROM ducknng_aio_collect((SELECT list_value(recv_aio) FROM pushpull_recv), 1000);

SET VARIABLE pushpull_drop = ducknng_aio_drop((SELECT recv_aio FROM pushpull_recv));
DROP TABLE pushpull_recv;
SET VARIABLE push_close = ducknng_close_socket(getvariable('push_socket')::UBIGINT);
SET VARIABLE pull_close = ducknng_close_socket(getvariable('pull_socket')::UBIGINT);
```

### Publish one raw message through `pub` / `sub`

```{sql eval=TRUE, collapse=FALSE, comment=""}
-- Open one publisher, then subscribe from a sub peer to all topics.
SET VARIABLE pub_socket = ducknng_open_socket('pub');
SET VARIABLE pub_listen_ok = ducknng_listen_socket(
  getvariable('pub_socket')::UBIGINT,
  'ipc:///tmp/ducknng_sql_pubsub_demo.ipc',
  134217728,
  0::UBIGINT
);
SET VARIABLE sub_socket = ducknng_open_socket('sub');
SET VARIABLE sub_subscribe_ok = ducknng_subscribe_socket(getvariable('sub_socket')::UBIGINT, from_hex(''));
SET VARIABLE sub_dial_ok = ducknng_dial_socket(
  getvariable('sub_socket')::UBIGINT,
  'ipc:///tmp/ducknng_sql_pubsub_demo.ipc',
  1000,
  0::UBIGINT
);

CREATE TEMP TABLE pubsub_recv AS
SELECT ducknng_recv_socket_raw_aio(getvariable('sub_socket')::UBIGINT, 1000) AS recv_aio;

SET VARIABLE pubsub_sent = ducknng_send_socket_raw(
  getvariable('pub_socket')::UBIGINT,
  from_hex('7075622d737562'),
  1000
);

SELECT getvariable('pubsub_sent')::BOOLEAN AS sent,
       ok,
       hex(frame) = '7075622D737562' AS got_payload
FROM ducknng_aio_collect((SELECT list_value(recv_aio) FROM pubsub_recv), 1000);

SET VARIABLE pubsub_drop = ducknng_aio_drop((SELECT recv_aio FROM pubsub_recv));
DROP TABLE pubsub_recv;
SET VARIABLE sub_close = ducknng_close_socket(getvariable('sub_socket')::UBIGINT);
SET VARIABLE pub_close = ducknng_close_socket(getvariable('pub_socket')::UBIGINT);
```

### Exchange one survey and one response through `surveyor` / `respondent`

```{sql eval=TRUE, collapse=FALSE, comment=""}
-- Open one respondent listener and one surveyor peer.
SET VARIABLE respondent_socket = ducknng_open_socket('respondent');
SET VARIABLE respondent_listen_ok = ducknng_listen_socket(
  getvariable('respondent_socket')::UBIGINT,
  'ipc:///tmp/ducknng_sql_survey_demo.ipc',
  134217728,
  0::UBIGINT
);
SET VARIABLE surveyor_socket = ducknng_open_socket('surveyor');
SET VARIABLE surveyor_dial_ok = ducknng_dial_socket(
  getvariable('surveyor_socket')::UBIGINT,
  'ipc:///tmp/ducknng_sql_survey_demo.ipc',
  1000,
  0::UBIGINT
);

-- Receive the survey on the respondent side first.
CREATE TEMP TABLE respondent_recv AS
SELECT ducknng_recv_socket_raw_aio(getvariable('respondent_socket')::UBIGINT, 1000) AS recv_aio;

SET VARIABLE survey_sent = ducknng_send_socket_raw(
  getvariable('surveyor_socket')::UBIGINT,
  from_hex('737572766579'),
  1000
);

SELECT getvariable('survey_sent')::BOOLEAN AS sent_survey,
       ok,
       hex(frame) = '737572766579' AS got_survey
FROM ducknng_aio_collect((SELECT list_value(recv_aio) FROM respondent_recv), 1000);

-- Then receive the respondent reply back on the surveyor side.
CREATE TEMP TABLE surveyor_recv AS
SELECT ducknng_recv_socket_raw_aio(getvariable('surveyor_socket')::UBIGINT, 1000) AS recv_aio;

SET VARIABLE response_sent = ducknng_send_socket_raw(
  getvariable('respondent_socket')::UBIGINT,
  from_hex('726573706f6e7365'),
  1000
);

SELECT getvariable('response_sent')::BOOLEAN AS sent_response,
       ok,
       hex(frame) = '726573706F6E7365' AS got_response
FROM ducknng_aio_collect((SELECT list_value(recv_aio) FROM surveyor_recv), 1000);

SET VARIABLE respondent_drop = ducknng_aio_drop((SELECT recv_aio FROM respondent_recv));
SET VARIABLE surveyor_drop = ducknng_aio_drop((SELECT recv_aio FROM surveyor_recv));
DROP TABLE respondent_recv;
DROP TABLE surveyor_recv;
SET VARIABLE surveyor_close = ducknng_close_socket(getvariable('surveyor_socket')::UBIGINT);
SET VARIABLE respondent_close = ducknng_close_socket(getvariable('respondent_socket')::UBIGINT);
```

### Broadcast one raw message through a three-peer `bus` mesh

```{sql eval=TRUE, collapse=FALSE, comment=""}
-- Open three bus peers. Every peer in a bus sends to and receives from every
-- other peer, so a single send from peer A is delivered to both B and C.
SET VARIABLE bus_a = ducknng_open_socket('bus');
SET VARIABLE bus_b = ducknng_open_socket('bus');
SET VARIABLE bus_c = ducknng_open_socket('bus');

-- A listens, B and C dial A. That gives every peer a direct edge to A; bus
-- forwards the broadcast across the mesh.
SET VARIABLE bus_a_listen_ok = ducknng_listen_socket(
  getvariable('bus_a')::UBIGINT,
  'ipc:///tmp/ducknng_sql_bus_demo.ipc',
  134217728,
  0::UBIGINT
);
SET VARIABLE bus_b_dial_ok = ducknng_dial_socket(
  getvariable('bus_b')::UBIGINT,
  'ipc:///tmp/ducknng_sql_bus_demo.ipc',
  1000,
  0::UBIGINT
);
SET VARIABLE bus_c_dial_ok = ducknng_dial_socket(
  getvariable('bus_c')::UBIGINT,
  'ipc:///tmp/ducknng_sql_bus_demo.ipc',
  1000,
  0::UBIGINT
);

-- Arm one async receive on each non-sending peer before A broadcasts.
CREATE TEMP TABLE bus_recv_b AS
SELECT ducknng_recv_socket_raw_aio(getvariable('bus_b')::UBIGINT, 1000) AS recv_aio;
CREATE TEMP TABLE bus_recv_c AS
SELECT ducknng_recv_socket_raw_aio(getvariable('bus_c')::UBIGINT, 1000) AS recv_aio;

SET VARIABLE bus_sent = ducknng_send_socket_raw(
  getvariable('bus_a')::UBIGINT,
  from_hex('6275732d62726f616463617374'),
  1000
);

SELECT getvariable('bus_sent')::BOOLEAN AS sent,
       ok,
       hex(frame) = '6275732D62726F616463617374' AS got_payload
FROM ducknng_aio_collect((SELECT list_value(recv_aio) FROM bus_recv_b), 1000);

SELECT ok,
       hex(frame) = '6275732D62726F616463617374' AS got_payload
FROM ducknng_aio_collect((SELECT list_value(recv_aio) FROM bus_recv_c), 1000);

SET VARIABLE bus_drop_b = ducknng_aio_drop((SELECT recv_aio FROM bus_recv_b));
SET VARIABLE bus_drop_c = ducknng_aio_drop((SELECT recv_aio FROM bus_recv_c));
DROP TABLE bus_recv_b;
DROP TABLE bus_recv_c;
SET VARIABLE bus_close_c = ducknng_close_socket(getvariable('bus_c')::UBIGINT);
SET VARIABLE bus_close_b = ducknng_close_socket(getvariable('bus_b')::UBIGINT);
SET VARIABLE bus_close_a = ducknng_close_socket(getvariable('bus_a')::UBIGINT);
```

### Launch raw requests asynchronously and collect the reply frames later

The stable async contract is raw-result-first. Framed RPC aio helpers collect raw reply frames through `ducknng_aio_collect(...)`, low-level HTTP aio helpers collect HTTP-shaped rows through `ducknng_ncurl_aio_collect(...)`, and callers explicitly decode frames or bodies afterward. URL-launched request aio helpers keep the same operation-oriented routing as the synchronous helpers: NNG URLs use NNG's background dial path, HTTP/HTTPS URLs use NNG's asynchronous HTTP client path, and both still terminate as one collected reply frame or one terminal error. Structured async wrappers may be added later as conveniences, but they remain additive; aio handles represent one pending operation, not a background job or streaming protocol.

```{sql eval=TRUE, collapse=FALSE, comment=""}
-- Start a local listener that the aio requests will target.
SELECT ducknng_start_server(
  'sql_aio_demo',                -- service name
  'ipc:///tmp/ducknng_sql_aio_demo.ipc', -- listen URL
  1,                             -- REP contexts
  134217728,                     -- recv_max_bytes
  300000,                        -- session_idle_ms
  0                              -- tls_config_id (0 means plaintext)
);

-- Launch two raw request/reply futures and keep their aio ids in one temp row.
-- timeout_ms bounds the pending network operation itself; it is separate from the later wait_ms passed to ducknng_aio_collect().
-- The hex literal below is the built-in manifest request frame in the current wire format.
CREATE TEMP TABLE aio_demo AS
SELECT
  ducknng_request_raw_aio(
    'ipc:///tmp/ducknng_sql_aio_demo.ipc', -- url
    from_hex('01000000000000000000000000000000000000000000'), -- manifest request frame
    1000,                                  -- timeout_ms
    0::UBIGINT                             -- tls_config_id
  ) AS aio1,
  ducknng_request_raw_aio(
    'ipc:///tmp/ducknng_sql_aio_demo.ipc', -- url
    from_hex('01000000000000000000000000000000000000000000'), -- manifest request frame
    1000,                                  -- timeout_ms
    0::UBIGINT                             -- tls_config_id
  ) AS aio2;

-- Confirm that both aio launches returned runtime handle ids.
SELECT aio1 > 0 AS aio1_started, aio2 > aio1 AS aio2_started_after_aio1
FROM aio_demo;

-- Wait for terminal results and return the full reply frames for later decoding.
SELECT aio_id, ok, octet_length(frame) > 0 AS has_frame
FROM ducknng_aio_collect((SELECT list_value(aio1, aio2) FROM aio_demo), 1000)
ORDER BY aio_id;

-- A collected aio is terminal, so ready() stays true until the handle is dropped.
SELECT ducknng_aio_ready(aio1) AS aio1_ready, ducknng_aio_ready(aio2) AS aio2_ready
FROM aio_demo;

-- Drop the terminal aio handles explicitly because SQL has no destructor hook.
SELECT ducknng_aio_drop(aio1) AND ducknng_aio_drop(aio2) AS dropped
FROM aio_demo;

-- Remove the temp state and stop the demo listener.
DROP TABLE aio_demo;
SELECT ducknng_stop_server('sql_aio_demo');
```

### Launch unary RPC calls asynchronously and decode the replies later

These helpers sit above the same request/reply aio substrate as `ducknng_request_raw_aio(...)`, but they build the manifest and exec request frames for you. They still collect raw reply frames later, so decoding remains explicit and honest.

```{sql eval=TRUE, collapse=FALSE, comment=""}
-- Start a local listener for the async RPC wrapper demo.
SELECT ducknng_start_server(
  'sql_rpc_aio_demo',            -- service name
  'ipc:///tmp/ducknng_sql_rpc_aio_demo.ipc', -- listen URL
  1,                             -- REP contexts
  134217728,                     -- recv_max_bytes
  300000,                        -- session_idle_ms
  0                              -- tls_config_id (0 means plaintext)
);

-- Register exec so the later async exec wrapper has a real method to call.
SELECT ducknng_register_exec_method();

-- Launch one manifest request and one metadata-only exec request asynchronously.
SET VARIABLE manifest_aio = ducknng_get_rpc_manifest_raw_aio(
  'ipc:///tmp/ducknng_sql_rpc_aio_demo.ipc', -- url
  1000,                                      -- timeout_ms
  0::UBIGINT                                 -- tls_config_id
);

SET VARIABLE exec_aio = ducknng_run_rpc_raw_aio(
  'ipc:///tmp/ducknng_sql_rpc_aio_demo.ipc', -- url
  'CREATE TABLE IF NOT EXISTS rpc_aio_demo_t(i INTEGER)', -- sql
  1000,                                      -- timeout_ms
  0::UBIGINT                                 -- tls_config_id
);

SELECT getvariable('manifest_aio') > 0 AS manifest_aio_started,
       getvariable('exec_aio') > getvariable('manifest_aio') AS exec_aio_started_after_manifest
;

-- Collect both terminal results. The collected values are still raw frames.
CREATE TEMP TABLE rpc_aio_collect AS
SELECT *
FROM ducknng_aio_collect(list_value(getvariable('manifest_aio'), getvariable('exec_aio')), 1000);

-- Store the collected raw frames explicitly before decoding them.
SET VARIABLE manifest_frame = (
  SELECT frame
  FROM rpc_aio_collect
  WHERE aio_id = getvariable('manifest_aio')
);

SET VARIABLE exec_frame = (
  SELECT frame
  FROM rpc_aio_collect
  WHERE aio_id = getvariable('exec_aio')
);

-- Decode the manifest reply frame explicitly after collection.
SELECT ok, type_name, name, position('"name":"exec"' IN payload_text) > 0 AS has_exec
FROM ducknng_decode_frame(getvariable('manifest_frame'));

-- Decode the async exec reply frame the same way.
SELECT ok, type_name, name
FROM ducknng_decode_frame(getvariable('exec_frame'));

-- Drop the terminal aio handles, remove the temp state, and stop the demo server.
SELECT ducknng_aio_drop(getvariable('manifest_aio')) AND ducknng_aio_drop(getvariable('exec_aio')) AS dropped;
DROP TABLE rpc_aio_collect;
SELECT ducknng_stop_server('sql_rpc_aio_demo');
```

### Open, fetch, and close a query session explicitly

```{sql eval=TRUE, collapse=FALSE, comment=""}
-- Start a listener that will own the server-side query session.
SELECT ducknng_start_server(
  'sql_session_demo',            -- service name
  'ipc:///tmp/ducknng_sql_session_demo.ipc', -- listen URL
  1,                             -- REP contexts
  134217728,                     -- recv_max_bytes
  300000,                        -- session_idle_ms
  0                              -- tls_config_id (0 means plaintext)
);

-- Open one server-owned query session.
-- batch_rows = 0 and batch_bytes = 0 mean "use the server defaults" for this demo.
-- Keep both session_id and session_token: the token is the bearer capability
-- required by fetch, close, and cancel.
CREATE TEMP TABLE session_open AS
SELECT *
FROM ducknng_open_query(
  'ipc:///tmp/ducknng_sql_session_demo.ipc', -- url
  'SELECT 1 AS id UNION ALL SELECT 2 AS id ORDER BY id', -- SQL text to run remotely
  0::UBIGINT,                                -- batch_rows
  0::UBIGINT,                                -- batch_bytes
  0::UBIGINT                                 -- tls_config_id
);

SET VARIABLE session_id = (SELECT session_id FROM session_open);
SET VARIABLE session_token = (SELECT session_token FROM session_open);

SELECT getvariable('session_id') AS opened_session_id,
       length(getvariable('session_token')::VARCHAR) AS session_token_chars,
       idle_timeout_ms AS effective_idle_timeout_ms
FROM session_open;

-- Use the table helper when you want one fetch batch decoded directly as rows.
SELECT *
FROM ducknng_fetch_query_table(
  'ipc:///tmp/ducknng_sql_session_demo.ipc', -- url
  getvariable('session_id')::UBIGINT,        -- session_id returned by open_query
  getvariable('session_token')::VARCHAR,     -- session_token returned by open_query
  0::UBIGINT,                                -- batch_rows override
  0::UBIGINT,                                -- batch_bytes override
  0::UBIGINT                                 -- tls_config_id
);

-- The raw fetch helper still exposes control_json, payload, and end_of_stream
-- explicitly. After the table helper consumes the first batch, the next fetch
-- is already the exhausted control reply.
SELECT ok, session_id, state, payload IS NULL AS no_payload, end_of_stream
FROM ducknng_fetch_query(
  'ipc:///tmp/ducknng_sql_session_demo.ipc', -- url
  getvariable('session_id')::UBIGINT,        -- session_id returned by open_query
  getvariable('session_token')::VARCHAR,     -- session_token returned by open_query
  0::UBIGINT,                                -- batch_rows override
  0::UBIGINT,                                -- batch_bytes override
  0::UBIGINT                                 -- tls_config_id
);

-- Close the exhausted session explicitly.
SELECT ok, session_id, state
FROM ducknng_close_query(
  'ipc:///tmp/ducknng_sql_session_demo.ipc', -- url
  getvariable('session_id')::UBIGINT,        -- session_id returned by open_query
  getvariable('session_token')::VARCHAR,     -- session_token returned by open_query
  0::UBIGINT                                 -- tls_config_id
);

-- Stop the demo listener after the session is closed.
SELECT ducknng_stop_server('sql_session_demo');
```

### Launch the session lifecycle asynchronously and collect raw frames

```{sql eval=TRUE, collapse=FALSE, comment=""}
-- Start a listener for the async session lifecycle demo.
SELECT ducknng_start_server(
  'sql_session_aio_demo',
  'ipc:///tmp/ducknng_sql_session_aio_demo.ipc',
  1,
  134217728,
  300000,
  0
);

-- Launch query_open asynchronously and collect the raw reply frame.
SET VARIABLE session_open_aio = ducknng_open_query_raw_aio(
  'ipc:///tmp/ducknng_sql_session_aio_demo.ipc',
  'SELECT 1 AS id UNION ALL SELECT 2 AS id ORDER BY id',
  0::UBIGINT,
  0::UBIGINT,
  1000,
  0::UBIGINT
);

CREATE TEMP TABLE session_open_aio_collect AS
SELECT *
FROM ducknng_aio_collect(list_value(getvariable('session_open_aio')::UBIGINT), 1000);

SET VARIABLE session_open_frame = (
  SELECT frame FROM session_open_aio_collect
);

SELECT ok, type_name, name, position('"session_id":' IN payload_text) > 0 AS has_session_id
FROM ducknng_decode_frame(getvariable('session_open_frame')::BLOB);

SET VARIABLE session_aio_id = (
  SELECT json_extract(payload_text::JSON, '$.session_id')::UBIGINT
  FROM ducknng_decode_frame(getvariable('session_open_frame')::BLOB)
);

SET VARIABLE session_aio_token = (
  SELECT json_extract_string(payload_text::JSON, '$.session_token')
  FROM ducknng_decode_frame(getvariable('session_open_frame')::BLOB)
);

SELECT ducknng_aio_drop(getvariable('session_open_aio')::UBIGINT) AS dropped_open_aio;
DROP TABLE session_open_aio_collect;

-- Launch fetch asynchronously and decode the collected raw frame explicitly.
SET VARIABLE session_fetch_aio = ducknng_fetch_query_raw_aio(
  'ipc:///tmp/ducknng_sql_session_aio_demo.ipc',
  getvariable('session_aio_id')::UBIGINT,
  getvariable('session_aio_token')::VARCHAR,
  0::UBIGINT,
  0::UBIGINT,
  1000,
  0::UBIGINT
);

CREATE TEMP TABLE session_fetch_aio_collect AS
SELECT *
FROM ducknng_aio_collect(list_value(getvariable('session_fetch_aio')::UBIGINT), 1000);

SET VARIABLE session_fetch_frame = (
  SELECT frame FROM session_fetch_aio_collect
);

SELECT ok, type_name, name, octet_length(payload) > 0 AS has_payload
FROM ducknng_decode_frame(getvariable('session_fetch_frame')::BLOB);

SELECT ducknng_aio_drop(getvariable('session_fetch_aio')::UBIGINT) AS dropped_fetch_aio;
DROP TABLE session_fetch_aio_collect;

-- Launch close asynchronously and decode the control reply frame.
SET VARIABLE session_close_aio = ducknng_close_query_raw_aio(
  'ipc:///tmp/ducknng_sql_session_aio_demo.ipc',
  getvariable('session_aio_id')::UBIGINT,
  getvariable('session_aio_token')::VARCHAR,
  1000,
  0::UBIGINT
);

CREATE TEMP TABLE session_close_aio_collect AS
SELECT *
FROM ducknng_aio_collect(list_value(getvariable('session_close_aio')::UBIGINT), 1000);

SET VARIABLE session_close_frame = (
  SELECT frame FROM session_close_aio_collect
);

SELECT ok, type_name, name, position('"state":"closed"' IN payload_text) > 0 AS is_closed
FROM ducknng_decode_frame(getvariable('session_close_frame')::BLOB);

SELECT ducknng_aio_drop(getvariable('session_close_aio')::UBIGINT) AS dropped_close_aio;
DROP TABLE session_close_aio_collect;

-- Open another session so the cancel aio helper has a live target.
CREATE TEMP TABLE session_cancel_open AS
SELECT *
FROM ducknng_open_query(
  'ipc:///tmp/ducknng_sql_session_aio_demo.ipc',
  'SELECT 99 AS id',
  0::UBIGINT,
  0::UBIGINT,
  0::UBIGINT
);

SET VARIABLE session_cancel_id = (
  SELECT session_id FROM session_cancel_open
);

SET VARIABLE session_cancel_token = (
  SELECT session_token FROM session_cancel_open
);

SET VARIABLE session_cancel_aio = ducknng_cancel_query_raw_aio(
  'ipc:///tmp/ducknng_sql_session_aio_demo.ipc',
  getvariable('session_cancel_id')::UBIGINT,
  getvariable('session_cancel_token')::VARCHAR,
  1000,
  0::UBIGINT
);

CREATE TEMP TABLE session_cancel_aio_collect AS
SELECT *
FROM ducknng_aio_collect(list_value(getvariable('session_cancel_aio')::UBIGINT), 1000);

SET VARIABLE session_cancel_frame = (
  SELECT frame FROM session_cancel_aio_collect
);

SELECT ok, type_name, name, position('"state":"cancelled"' IN payload_text) > 0 AS is_cancelled
FROM ducknng_decode_frame(getvariable('session_cancel_frame')::BLOB);

SELECT ducknng_aio_drop(getvariable('session_cancel_aio')::UBIGINT) AS dropped_cancel_aio;
DROP TABLE session_cancel_open;
DROP TABLE session_cancel_aio_collect;

SELECT ducknng_stop_server('sql_session_aio_demo');
```

### `tls+tcp://` with a self-signed development TLS config

```{sql eval=TRUE, collapse=FALSE, comment=""}
-- Generate a self-signed loopback certificate and keep it as a runtime TLS handle.
SELECT ducknng_self_signed_tls_config('127.0.0.1', 365, 0);

-- Start a TLS listener using the generated config handle.
SELECT ducknng_start_server(
  'tls_demo_self',               -- service name
  'tls+tcp://127.0.0.1:45453',   -- listen URL
  1,                             -- REP contexts
  134217728,                     -- recv_max_bytes
  300000,                        -- session_idle_ms
  1                              -- tls_config_id returned above
);

-- Send the built-in manifest request frame over TLS and decode the reply.
SELECT ok, type_name, name, position('"name":"exec"' IN payload_text) > 0
FROM ducknng_decode_frame(
  ducknng_request_raw(
    'tls+tcp://127.0.0.1:45453',                -- url
    from_hex('01000000000000000000000000000000000000000000'), -- manifest request frame
    1000,                                       -- timeout_ms
    1                                           -- tls_config_id
  )
);

-- Clean up the TLS demo server and config handle.
SELECT ducknng_stop_server('tls_demo_self');
SELECT ducknng_drop_tls_config(1);
```

For mTLS, create the same kind of TLS handle with `auth_mode = 2`. On listeners this requires the peer to present a certificate trusted by the configured CA; on clients it requires server verification. The dispatcher derives the current caller identity from the first verified peer certificate SAN as `tls:san:<value>`, falling back to the common name as `tls:cn:<common-name>`. Sessions opened over mTLS are still controlled by `session_token`, but they are also bound to that verified peer identity. `ducknng_set_tls_peer_allowlist(...)` sets an exact identity allowlist copied into future services, and `ducknng_set_service_peer_allowlist(...)` changes the allowlist dynamically for a running service. NNG listeners use NNG's `NNG_PIPE_EV_ADD_PRE` pipe notification to close non-admitted new pipes before they are added to the socket; HTTP/HTTPS returns HTTP `403` before RPC dispatch. `ducknng_list_servers()` exposes `active_pipes`, `max_active_pipes`, `inflight_requests`, `max_inflight_requests`, `max_sessions_per_peer_identity`, `tls_enabled`, `tls_auth_mode`, `peer_identity_required`, `peer_allowlist_active`, `ip_allowlist_active`, `sql_authorizer_active`, and the corresponding counts so deployments can distinguish current NNG membership, TLS without client verification, mTLS, allowlisted mTLS, IP-gated services, and services with SQL authorization callbacks. Individual registry-backed methods can also require verified peer identity; for example, `ducknng_set_method_auth('manifest', true)` protects manifest discovery without unregistering the method.

The built-in mTLS, peer-identity allowlist, and IP/CIDR allowlist checks are the fast C path for common denials. They run before the flexible SQL callback and, for NNG transports, can reject new pipes at `NNG_PIPE_EV_ADD_PRE` without entering DuckDB. Basic service resource limits also start in C: `ducknng_set_service_limits(name, max_open_sessions)` caps concurrently open query sessions for a service, the three-argument form also caps simultaneously active NNG protocol-socket pipes at `ADD_PRE`, and the four-argument form also caps concurrent in-flight requests before SQL authorizers and RPC dispatch, and the five-argument form also caps concurrently open query sessions per verified peer identity; `0` means unlimited for any cap. Those are the stable built-in quota owners today: the service itself and, when present, the verified peer identity. The optional SQL-authorizer `principal` column is deployment policy/audit metadata, not yet durable session ownership metadata for built-in quotas. For policy that depends on tables, tenants, method names, headers, or deployment-specific rules, install a service-level SQL authorizer with `ducknng_set_service_authorizer(name, authorizer_sql)`. The callback sees one row from `ducknng_auth_context()` while it runs and must return exactly one row with a Boolean `allow` column; optional columns are `status`, `reason`, `principal`, `claims_json`, and `cache_ttl_ms`. `NULL` or an empty string clears the SQL authorizer. This callback is intentionally evaluated at the request/dispatch boundary, not inside NNG's low-level pipe callback. It runs through the service-owned DuckDB execution lane, so keep it short and side-effect-light: prefer table/view lookups, avoid recursive `ducknng_*` client calls to the same service, avoid stopping the service from its own callback, and do not use it for long-running work. That limitation avoids deadlocks while keeping one uniform policy interface for NNG, HTTP/HTTPS framed RPC, and the planned broader HTTP server framework.

### `tls+tcp://` from file-backed certificate material

```{sql eval=TRUE, collapse=FALSE, comment=""}
-- Register a file-backed TLS config using committed loopback test certificates.
SELECT ducknng_tls_config_from_files(
  'test/certs/loopback-cert-key.pem', -- cert_key_file
  'test/certs/loopback-ca.pem',       -- ca_file
  NULL,                               -- password
  0                                   -- auth_mode
);

-- Start a TLS listener with that file-backed config.
SELECT ducknng_start_server(
  'tls_demo_files',              -- service name
  'tls+tcp://127.0.0.1:45454',   -- listen URL
  1,                             -- REP contexts
  134217728,                     -- recv_max_bytes
  300000,                        -- session_idle_ms
  1                              -- tls_config_id returned above
);

-- Inspect the manifest reply over TLS using the same config handle on the client side.
SELECT ok, type_name, name, position('"name":"exec"' IN payload_text) > 0
FROM ducknng_decode_frame(
  ducknng_request_raw(
    'tls+tcp://127.0.0.1:45454',                -- url
    from_hex('01000000000000000000000000000000000000000000'), -- manifest request frame
    1000,                                       -- timeout_ms
    1                                           -- tls_config_id
  )
);

-- Clean up the file-backed TLS demo server and config handle.
SELECT ducknng_stop_server('tls_demo_files');
SELECT ducknng_drop_tls_config(1);
```

### `ws://` and `wss://` as NNG transports

`ws://` and `wss://` are NNG transport schemes, not the HTTP carrier described in `docs/http.md`. They use `ducknng_start_server(...)`, the same framed manifest/RPC/session helpers, and the same TLS handle model as the other NNG transports.

```{sql eval=TRUE, collapse=FALSE, comment=""}
-- Plain WebSocket transport through the NNG service layer.
SELECT ducknng_start_server(
  'ws_demo',
  'ws://127.0.0.1:45455/_ducknng',
  1,
  134217728,
  300000,
  0::UBIGINT
);

SELECT ok, type_name, name
FROM ducknng_decode_frame(
  ducknng_request_raw(
    'ws://127.0.0.1:45455/_ducknng',
    from_hex('01000000000000000000000000000000000000000000'),
    1000,
    0::UBIGINT
  )
);

SELECT ducknng_stop_server('ws_demo');

-- Secure WebSocket transport uses the same reusable TLS handle model.
SELECT ducknng_self_signed_tls_config('127.0.0.1', 365, 0);

SELECT ducknng_start_server(
  'wss_demo',
  'wss://127.0.0.1:45456/_ducknng',
  1,
  134217728,
  300000,
  1::UBIGINT
);

SELECT ok, type_name, name
FROM ducknng_decode_frame(
  ducknng_request_raw(
    'wss://127.0.0.1:45456/_ducknng',
    from_hex('01000000000000000000000000000000000000000000'),
    1000,
    1::UBIGINT
  )
);

SELECT ducknng_stop_server('wss_demo');
SELECT ducknng_drop_tls_config(1);
```

### REQ/REP `EXEC` via `nanonext` as an interop example

This example is easier to follow as a short sequence: define the frame helpers, define the retry helpers, start a local server, run a few remote `exec` requests, inspect the same table locally, and then clean up.

#### Frame encoding and decoding helpers

```{r exec-example-frame-helpers}
# Little-endian helpers for the versioned ducknng RPC envelope.
write_u32le <- function(x) {
  writeBin(as.integer(x), raw(), size = 4L, endian = "little")
}

write_u64le <- function(x) {
  x <- as.double(x)
  c(write_u32le(x %% 2^32), write_u32le(floor(x / 2^32)))
}

read_u32le_bytes <- function(x) {
  sum(as.double(as.integer(x)) * 256^(0:3))
}

read_u64le_bytes <- function(x) {
  read_u32le_bytes(x[1:4]) + 2^32 * read_u32le_bytes(x[5:8])
}

read_u32le <- function(buf, offset) {
  read_u32le_bytes(buf[offset + 0:3])
}

read_u64le <- function(buf, offset) {
  read_u64le_bytes(buf[offset + 0:7])
}

# Encode the Arrow IPC payload carried by the built-in exec RPC method.
encode_ducknng_exec_payload <- function(sql, want_result = FALSE) {
  raw_con <- rawConnection(raw(), open = "r+")
  on.exit(close(raw_con), add = TRUE)

  nanoarrow::write_nanoarrow(
    data.frame(sql = sql, want_result = want_result),
    raw_con
  )

  rawConnectionValue(raw_con)
}

# Build a version 1 RPC call frame for the named method.
encode_ducknng_call_frame <- function(method, payload, flags = 0L) {
  method_name <- charToRaw(method)

  c(
    as.raw(1L),
    as.raw(1L),
    write_u32le(flags),
    write_u32le(length(method_name)),
    write_u32le(0),
    write_u64le(length(payload)),
    method_name,
    payload
  )
}

encode_ducknng_exec_request <- function(sql, want_result = FALSE) {
  encode_ducknng_call_frame(
    method = "exec",
    payload = encode_ducknng_exec_payload(sql, want_result = want_result)
  )
}

# Decode a reply envelope and parse its Arrow IPC payload when present.
decode_ducknng_exec_reply <- function(buf) {
  header_size <- 23L

  if (!is.raw(buf) || length(buf) < (header_size - 1L)) {
    stop("ducknng reply frame was empty or truncated", call. = FALSE)
  }

  name_len <- read_u32le(buf, 7)
  error_len <- read_u32le(buf, 11)
  payload_len <- read_u64le(buf, 15)

  name_start <- header_size
  error_start <- name_start + name_len
  payload_start <- error_start + error_len
  payload_end <- payload_start + payload_len - 1L

  payload <- if (payload_len > 0) buf[payload_start:payload_end] else raw()

  list(
    version = as.integer(buf[1]),
    type = as.integer(buf[2]),
    flags = read_u32le(buf, 3),
    method = rawToChar(buf[name_start:(error_start - 1L)]),
    error = if (error_len > 0) rawToChar(buf[error_start:(payload_start - 1L)]) else "",
    payload = payload,
    data = if (payload_len > 0) as.data.frame(nanoarrow::read_nanoarrow(payload)) else NULL
  )
}
```

#### REQ/REP retry helpers

```{r exec-example-retry-helpers}
# Send and receive with the same retry style used in mangoro examples.
send_ducknng_frame <- function(socket, frame, timeout_ms = 1000L, max_attempts = 20L) {
  attempt <- 1L
  send_result <- nanonext::send(socket, frame, mode = "raw", block = timeout_ms)

  while (nanonext::is_error_value(send_result) && attempt < max_attempts) {
    Sys.sleep(0.25)
    send_result <- nanonext::send(socket, frame, mode = "raw", block = timeout_ms)
    attempt <- attempt + 1L
  }

  send_result
}

recv_ducknng_frame <- function(socket, timeout_ms = 1000L, max_attempts = 20L) {
  attempt <- 1L
  response <- nanonext::recv(socket, mode = "raw", block = timeout_ms)

  while (
    (
      nanonext::is_error_value(response) ||
      !is.raw(response) ||
      length(response) < 22L
    ) &&
    attempt < max_attempts
  ) {
    Sys.sleep(0.25)
    response <- nanonext::recv(socket, mode = "raw", block = timeout_ms)
    attempt <- attempt + 1L
  }

  response
}

run_ducknng_req <- function(socket, frame, timeout_ms = 1000L) {
  send_ducknng_frame(socket, frame, timeout_ms = timeout_ms)
  decode_ducknng_exec_reply(
    recv_ducknng_frame(socket, timeout_ms = timeout_ms)
  )
}

# Wait for an IPC listener path to appear before dialing it.
wait_for_ducknng_listener <- function(path, timeout_secs = 10, interval_secs = 0.1) {
  deadline <- Sys.time() + timeout_secs

  while (Sys.time() < deadline) {
    if (file.exists(path)) {
      return(invisible(TRUE))
    }
    Sys.sleep(interval_secs)
  }

  stop("ducknng IPC listener did not become ready in time", call. = FALSE)
}
```

#### Start a local ducknng server and dial a req socket

```{r exec-example-setup}
# Start a ducknng server in this R session, then talk to it over a req socket.
ext_path <- normalizePath("build/release/ducknng.duckdb_extension")
ipc_path <- "/tmp/ducknng_readme_exec.ipc"
unlink(ipc_path)
ipc_url <- paste0("ipc://", ipc_path)

db_driver <- duckdb::duckdb(
  dbdir = ":memory:",
  config = list(allow_unsigned_extensions = "true")
)
db_con <- DBI::dbConnect(db_driver)

DBI::dbGetQuery(db_con, "SELECT 42 AS ok")
DBI::dbExecute(db_con, sprintf("LOAD '%s'", ext_path))
DBI::dbGetQuery(
  db_con,
  sprintf(
    "SELECT ducknng_start_server('sql_exec', '%s', 1, 134217728, 300000, 0::UBIGINT)",
    ipc_url
  )
)
DBI::dbGetQuery(db_con, "SELECT ducknng_register_exec_method()")

wait_for_ducknng_listener(ipc_path)
req <- nanonext::socket("req", dial = ipc_url)
```

#### Create and insert rows remotely

```{r exec-example-mutate}
# Create a table remotely and inspect the metadata reply.
create_reply <- run_ducknng_req(
  req,
  encode_ducknng_exec_request("CREATE TABLE ducknng_exec_demo(i INTEGER)")
)
create_reply$data

# Insert rows remotely and inspect the second metadata reply.
insert_reply <- run_ducknng_req(
  req,
  encode_ducknng_exec_request("INSERT INTO ducknng_exec_demo VALUES (42), (99)")
)
insert_reply$data
```

#### Query rows remotely with `want_result = TRUE`

```{r exec-example-query}
select_reply <- run_ducknng_req(
  req,
  encode_ducknng_exec_request(
    "SELECT i, i > 50 AS gt_50 FROM ducknng_exec_demo ORDER BY i",
    want_result = TRUE
  )
)
select_reply$data
```

#### Inspect the same table locally through DuckDB

```{r exec-example-local}
server_rows <- DBI::dbGetQuery(db_con, "SELECT * FROM ducknng_exec_demo ORDER BY i")
server_rows
```

#### Stop the server and clean up

```{r exec-example-cleanup}
DBI::dbGetQuery(db_con, "SELECT ducknng_stop_server('sql_exec')")
close(req)
DBI::dbDisconnect(db_con)
unlink(ipc_path)
```

## Deployment and admission policy

`ducknng` defaults to low-level, default-open transport behavior. The application or deployment decides where the trust boundary is. For anything beyond process-local, filesystem-local, or loopback-only development, set the boundary deliberately before exposing a service.

A strong direct-exposure baseline is:

1. use `https://`, `wss://`, or `tls+tcp://` rather than plaintext network carriers;
2. create the TLS handle with `auth_mode = 2` so peers must present a verified client certificate;
3. configure a private CA or otherwise narrow certificate issuance so the trust root is not broader than the application;
4. set exact peer-identity and/or IP/CIDR allowlists before exposure with `ducknng_set_tls_peer_allowlist(...)` and the optional `ip_allowlist_json` startup argument;
5. set basic service resource limits such as `ducknng_set_service_limits(name, max_open_sessions[, max_active_pipes[, max_inflight_requests[, max_sessions_per_peer_identity]]])` before exposing query sessions;
6. install a service-level SQL authorizer with `ducknng_set_service_authorizer(...)` when admission depends on deployment tables, tenants, method names, or HTTP metadata;
7. optionally update a running service dynamically with `ducknng_set_service_peer_allowlist(...)`, `ducknng_set_service_ip_allowlist(...)`, `ducknng_set_service_limits(...)`, and `ducknng_set_service_authorizer(...)`.

Free-form SQL execution is intentionally a deployment-owned capability. `ducknng` keeps its own generated SQL injection-safe, but it does not make arbitrary user SQL safe: `exec` and query sessions run DuckDB SQL as the service process on the configured DuckDB connection. Do not expose `exec` or unrestricted query sessions to callers who are not already authorized to run that SQL. Use these profiles as a starting point:

- **Local development / single-user IPC**: `inproc://`, loopback, or `ipc://` under a private directory; `exec` may be enabled for convenience inside that local boundary.
- **Trusted service mesh**: `tls+tcp://`, `wss://`, or `https://` with mTLS, exact peer allowlists, IP/CIDR allowlists where useful, and service limits; register `exec` with `ducknng_register_exec_method(true)` or leave it disabled.
- **Shared or semi-trusted clients**: prefer query sessions plus a short SQL authorizer; use deployment-level DuckDB/process controls for filesystem, extension loading, outbound network, and attachments.
- **Public/untrusted internet**: do not expose raw DuckDB SQL directly. Put `ducknng` behind a gateway that authenticates, rate-limits, and maps users to fixed application operations, or build a separate route layer that does not forward arbitrary SQL text.

When writing SQL authorizers, read request fields from `ducknng_auth_context()` and avoid string-concatenating untrusted HTTP headers, method names, or payload text into new SQL. Keep the callback short and side-effect-light.

The relevant identity strings currently come from verified TLS peer metadata and are SAN-first, CN-fallback: `tls:san:<value>` or `tls:cn:<common-name>`. Peer identity allowlists are exact-match. IP allowlists are parsed once into IPv4/IPv6 literal or CIDR rules such as `127.0.0.1/32`, `10.0.0.0/8`, or `::1/128`. `NULL` or an empty SQL string clears an allowlist and returns to default-open admission subject to TLS/method/session policy. An empty JSON array, `[]`, is active deny-all.

Example shape:

```sql
-- Set a default allowlist on the TLS handle before starting services with it.
SELECT ducknng_set_tls_peer_allowlist(
  1::UBIGINT,
  '["tls:san:client-a.example","tls:san:client-b.example"]'
);

-- Start with an IP/CIDR allowlist already installed before the listener accepts traffic.
SELECT ducknng_start_server(
  'sql_http',
  'http://127.0.0.1:18444/_ducknng',
  1,
  134217728,
  300000,
  0::UBIGINT,
  '["127.0.0.1/32"]'
);

-- Later, rotate or tighten admission for a running service.
SELECT ducknng_set_service_peer_allowlist(
  'sql_rpc',
  '["tls:san:client-b.example"]'
);
SELECT ducknng_set_service_ip_allowlist(
  'sql_http',
  '["127.0.0.1/32","10.0.0.0/8"]'
);

-- Set a basic service-side resource limit for query sessions.
SELECT ducknng_set_service_limits('sql_http', 16::UBIGINT);

-- Add a flexible SQL callback when policy depends on request context.
SELECT ducknng_set_service_authorizer(
  'sql_http',
  'SELECT remote_ip = ''127.0.0.1'' AND rpc_method = ''manifest'' AS allow,
          403 AS status,
          ''not authorized'' AS reason
   FROM ducknng_auth_context()'
);
```

Transport-specific deployment notes:

| Surface | Accepted schemes | TLS handle accepted on |
| --- | --- | --- |
| `ducknng_start_server(...)` | `inproc://`, `ipc://`, `tcp://`, `tls+tcp://`, `ws://`, `wss://`, `http://`, `https://` | `tls+tcp://`, `wss://`, `https://` |
| synchronous RPC/session helpers | NNG schemes plus `http://`, `https://` | `tls+tcp://`, `wss://`, `https://` |
| raw RPC AIO helpers | NNG schemes plus `http://`, `https://` | `tls+tcp://`, `wss://`, `https://` |
| generic socket APIs | `inproc://`, `ipc://`, `tcp://`, `tls+tcp://`, `ws://`, `wss://` | `tls+tcp://`, `wss://` |
| HTTP client helpers (`ducknng_ncurl(...)`, `ducknng_ncurl_aio(...)`, `ducknng_ncurl_table(...)`) | `http://`, `https://` | `https://` |

Supplying a non-zero TLS handle on a non-TLS URL is rejected rather than silently ignored. HTTP/HTTPS URLs are rejected by the generic NNG socket layer, and NNG URLs are rejected by HTTP-only helpers.

- `inproc://` is a same-process boundary. It is useful for embedded testing/composition, but it is not network authentication.
- `ipc://` relies on filesystem or named-pipe permissions. Avoid world-writable shared paths for multi-user deployments.
- plaintext `tcp://`, `ws://`, and `http://` do not provide cryptographic peer identity. Use them only on loopback, a private network you already trust, or behind a reverse proxy/firewall/VPN that performs admission itself.
- `http://` is still useful as a local or reverse-proxy upstream carrier. For public HTTP-style deployment, expose `https://` directly with mTLS or place `ducknng` behind a proxy such as Caddy, nginx, Envoy, or a cloud load balancer that terminates TLS and enforces authentication/rate limits before forwarding to loopback `http://` or `ipc://`.
- `tls+tcp://` and `wss://` are NNG protocol transports. They use NNG socket protocols and NNG pipes, not the HTTP carrier semantics.
- `https://` is the HTTP frame carrier over TLS. It shares the same framed RPC methods and admission policy, but HTTP status codes such as `403` can appear before RPC dispatch.

IP allowlisting is a useful lighter-weight gate when client certificates are too much operational overhead, but it is not the same as cryptographic client identity. HTTPS itself does not standardize an application IP allowlist; IP admission is normally implemented by the listener, firewall, reverse proxy, load balancer, or service mesh. `ducknng` now includes its own service-level IP/CIDR allowlist for direct listeners. For NNG protocol transports, it uses `NNG_OPT_REMADDR` from the NNG pipe and checks parsed binary CIDR rules during `ADD_PRE` and again at dispatch. For HTTP/HTTPS, it reads the connection remote address from the NNG supplemental HTTP connection and returns HTTP `403` before RPC dispatch when the address is not admitted. A front door such as Caddy, nginx, Envoy, cloud firewall/security groups, Kubernetes network policy, or a similar proxy remains useful for public deployments because it can combine IP/CIDR policy with rate limits, request logging, WAF rules, and ordinary web authentication before forwarding to loopback `http://`, `https://`, `ipc://`, or an internal NNG service.

For NNG transports, admission uses NNG's own pipe notification primitive: `nng_pipe_notify(..., NNG_PIPE_EV_ADD_PRE, ...)` lets `ducknng` inspect verified TLS pipe metadata and remote-address metadata and close a non-admitted pipe before it is added to the socket. For HTTP/HTTPS, NNG's supplemental HTTP API does not expose the same public protocol-socket pipe admission hook, so `ducknng` checks the same policy before RPC dispatch and returns HTTP `403` for non-admitted peers or remote addresses.

The same pipe-event family is useful beyond admission. NNG exposes `ADD_PRE`, `ADD_POST`, and `REM_POST` pipe events; nanonext surfaces related tools as `pipe_notify()`, `monitor()`, and `read_monitor()`. `ducknng` now records a bounded per-service NNG pipe event stream and exposes it with `ducknng_read_monitor(name, after_seq, max_events)`. It exposes ring/counter metadata with `ducknng_monitor_status(name)` and current active NNG pipes with `ducknng_list_pipes(name)`. The event stream includes event sequence, timestamp, service, transport, pipe id, admission result for `ADD_PRE`, denial reason when the fast C path has one, remote address, and verified peer identity when present; the active-pipe table gives the current pipe set for routing and membership decisions. These primitives are a natural foundation for telemetry, connection counts, connection churn, per-service pipe event streams, presence/worker membership, dynamic routing, backpressure-aware scheduling, mesh-style DuckDB service discovery, routing/forwarding demos, and other application-level event streams. HTTP/HTTPS framed RPC currently uses the same admission policy but does not expose NNG protocol-socket pipe events; a broader HTTP server framework can reuse the same request context and monitor concepts without changing the framed RPC wire protocol.

## Status

What is sealed and runnable in v1: NNG transports and socket patterns; first-class one-shot AIO across socket send/recv, ncurl HTTP, and unary RPC; framed RPC (`manifest`, opt-in `exec`, raw unary, query sessions); fast C admission (mTLS, exact peer-identity allowlists, IP/CIDR allowlists, service limits); SQL authorizer callbacks; bounded per-service pipe-event monitor and active-pipe snapshot; body codec layer with built-in providers and user-registered codec hooks. See `function_catalog/functions.md` for the exact surface and `NEWS.md` for landed changes.

Intentionally deferred:

- A broader nanonext-style HTTP route/server framework — the current HTTP server is the framed RPC mount.
- CSV/TSV/Parquet body parsers beyond the safe BLOB fallback.
- Full SQL-side decoding of session `fetch` Arrow batch BLOBs into a table-function path.
- Routing/forwarding and mesh scheduling as sealed APIs — the pipe monitor and active-pipe snapshot are the membership/telemetry foundation that comes first.

## References

- [NNG](https://nng.nanomsg.org/) for the underlying messaging library and transport family.
- [`nanonext`](https://github.com/r-lib/nanonext) for the main client/server ergonomics reference.
- [`mangoro`](https://github.com/sounkou-bioinfo/mangoro) for the thin-envelope + Arrow IPC RPC direction.
- [DuckDB C API](https://duckdb.org/docs/stable/clients/c/api) for the extension and SQL integration boundary.
- [Apache Arrow IPC](https://arrow.apache.org/docs/format/Columnar.html#serialization-and-interprocess-communication-ipc) for the tabular payload format.
