---
title: "ducknng bulk transport benchmark"
output:
  github_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "", echo = FALSE)
source(normalizePath("rpc_bulk_compare_support.R"))
knitr::opts_knit$set(root.dir = normalizePath(".."))

repetitions <- as.integer(Sys.getenv("DUCKNNG_BULK_REPETITIONS", unset = "5"))
rows <- ducknng_bench_parse_int_csv(
  Sys.getenv("DUCKNNG_BULK_ROWS", unset = ""),
  c(100000L, 1000000L, 10000000L)
)
db_path <- Sys.getenv(
  "DUCKNNG_BENCH_DB_PATH",
  unset = file.path(tempdir(), "ducknng_quack_tpch.duckdb")
)

```

This report is rendered by `make rpc_bulk_compare`. The default workload uses `DUCKNNG_BULK_ROWS=100000,1000000,10000000` and `DUCKNNG_BULK_REPETITIONS=5`; set those environment variables when you want a smoke render instead of the full benchmark.

It measures two things:

1. `ducknng` bulk row transfer over its RPC/session surface on `http`, `tcp`, `ipc`, and `ws`, in both `arrow_ipc_stream` and `quack_batch_v1` modes.
2. `quack` bulk row transfer over its published client/server surface.

`quack` is only compared on its own public client path. `ducknng` is compared across multiple transports because transport selection is part of its SQL-facing contract.

```{r run-benchmark, include=FALSE}
message(sprintf(
  "running bulk benchmark: repetitions=%d rows=%s db_path=%s",
  repetitions,
  paste(rows, collapse = ","),
  db_path
))
results <- ducknng_bench_run_bulk_compare(
  repetitions = repetitions,
  rows = rows,
  db_path = db_path
)

metadata_long <- data.frame(
  field = names(results$metadata),
  value = vapply(results$metadata[1, ], function(x) as.character(x[[1]]), character(1)),
  stringsAsFactors = FALSE
)

rpc_results <- results$rpc_results[order(
  results$rpc_results$transport,
  results$rpc_results$system,
  results$rpc_results$serialization_mode,
  results$rpc_results$rows
), ]
ratio_results <- results$http_vs_quack[order(results$http_vs_quack$rows), ]
```

## Machine and software details

```{r machine-details}
knitr::kable(metadata_long, col.names = c("Field", "Value"), row.names = FALSE)
```

## RPC bulk transfer results

These runs execute `SELECT * FROM lineitem LIMIT n` against the server side and validate the returned data by aggregate checksum.

```{r rpc-results}
knitr::kable(rpc_results, row.names = FALSE)
```

## HTTP ducknng vs Quack ratio

This is the apples-to-apples comparison on the common HTTP-facing path, showing both ducknng serializer modes against Quack's native HTTP path.

```{r http-ratio}
knitr::kable(ratio_results, row.names = FALSE)
```

## Notes

- The row benchmark uses a local TPC-H `lineitem` table generated through DuckDB's `tpch` extension when needed.
- `quack` is installed from `core_nightly` during the run if it is not already available.
