---
title: "ducknng upload lane vs text INSERT benchmark"
output:
  github_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "", echo = FALSE)
options(scipen = 999)
source(normalizePath("rpc_bulk_compare_support.R"))
source(normalizePath("rpc_upload_compare_support.R"))
knitr::opts_knit$set(root.dir = normalizePath(".."))

writer_counts <- ducknng_bench_parse_int_csv(
  Sys.getenv("DUCKNNG_UPLOAD_WRITERS", unset = ""),
  c(1L, 2L, 4L, 8L)
)
rows_per_writer <- as.integer(Sys.getenv("DUCKNNG_UPLOAD_ROWS", unset = "20000"))
repetitions <- as.integer(Sys.getenv("DUCKNNG_UPLOAD_REPETITIONS", unset = "3"))
transport <- Sys.getenv("DUCKNNG_UPLOAD_TRANSPORT", unset = "tcp")
ext_path <- Sys.getenv("DUCKNNG_BENCH_EXT_PATH", unset = ducknng_bench_default_upload_ext_path())
```

This report is rendered directly with `R -e "rmarkdown::render('bench/rpc_upload_compare.Rmd')"` (there is no Makefile target for it). The default workload is a quick smoke run: `DUCKNNG_UPLOAD_WRITERS=1,2,4,8`, `DUCKNNG_UPLOAD_ROWS=20000`, `DUCKNNG_UPLOAD_REPETITIONS=3`, `DUCKNNG_UPLOAD_TRANSPORT=tcp`. Scale up by raising `DUCKNNG_UPLOAD_ROWS` (rows written per writer per repetition) and/or `DUCKNNG_UPLOAD_REPETITIONS`.

## What is compared

Two client-to-server write paths. Both ship rows already materialized on the client — this is the point of the comparison, not "tell the server to make rows":

1. **text_insert** (baseline) — the client runs the row-generating query locally, formats every row as a SQL literal tuple, and sends one `INSERT INTO t VALUES (v1), (v2), ...` statement, with every row inlined as SQL text, through the existing `ducknng_run_rpc(...)` path.
2. **quack_upload** (treatment) — the client hands the *same* row-generating query to `ducknng_upload_table(url, source_query, target_table)`, which runs it locally and streams the result rows to the server over the quack upload lane (`upload_open` / `upload_append` / `upload_commit`, abort-and-rollback on error).

Both paths generate the identical deterministic dataset (`id INTEGER, seq BIGINT, value DOUBLE, label VARCHAR`) from the same seeded `range()`-based query, so their row counts and checksums must agree.

## Concurrency and correctness

For each writer count N in `writer_counts`, N concurrent writer processes (PSOCK cluster, one process per writer) each write to their **own** target table (`dst_w1`, `dst_w2`, ...), so single-table insert contention does not swamp the protocol comparison. Each (writer_mode x N) cell is repeated `repetitions` times; reported latency is the per-writer median and p95 across those repetitions.

After every single write, the benchmark re-reads the target table over the wire (`ducknng_query_rpc`) and checks its row count, `sum(value)`, and `count(DISTINCT id)` against a value computed independently in R (closed-form integer arithmetic, no DuckDB involved) for the expected dataset. Any mismatch raises an R error immediately and aborts the whole run — a fast-but-wrong upload is never reported as a clean result.

```{r run-benchmark, include=FALSE}
message(sprintf(
  "running upload benchmark: writers=%s rows_per_writer=%d repetitions=%d transport=%s",
  paste(writer_counts, collapse = ","), rows_per_writer, repetitions, transport
))
results_all <- ducknng_bench_run_upload_compare(
  writer_counts = writer_counts,
  writer_modes = c("text_insert", "quack_upload"),
  rows_per_writer = rows_per_writer,
  repetitions = repetitions,
  transport = transport,
  ext_path = ext_path
)

metadata_long <- data.frame(
  field = names(results_all$metadata),
  value = vapply(results_all$metadata[1, ], function(x) as.character(x[[1]]), character(1)),
  stringsAsFactors = FALSE
)
results <- results_all$results[order(results_all$results$writers, results_all$results$writer_mode), ]
size_ratio <- results_all$size_ratio
```

## Machine and software details

```{r machine-details}
knitr::kable(metadata_long, col.names = c("Field", "Value"), row.names = FALSE)
```

## Upload lane vs text INSERT results

`aggregate_rows_per_second` and `aggregate_mb_per_second` divide totals across all writers in the cell by `operation_seconds` (the slowest writer's total time), so cluster startup overhead does not dilute throughput. `total_bytes_shipped` is the actual client payload: total inlined-VALUES text bytes for `text_insert`, total `bytes_uploaded` for `quack_upload`. `median_latency_seconds` / `p95_latency_seconds` are per-writer, taken across `repetitions` repeats of a single write of `rows_per_writer` rows.

```{r upload-results}
knitr::kable(results, row.names = FALSE)
```

## Payload-size ratio (text / quack)

```{r size-ratio}
knitr::kable(size_ratio, row.names = FALSE)
```

## Correctness gate

Every cell above reports `correctness_gate = PASS`. That is not a formality: each individual write in every cell is independently re-read over the wire and checked against an analytically computed row count and checksum before the next repetition runs, and the benchmark function throws (halting the whole render) on the first mismatch it finds.

## Notes

- Dataset: each writer generates `rows_per_writer` deterministic rows (`id INTEGER, seq BIGINT, value DOUBLE, label VARCHAR`) from a seeded `range()` query; see `ducknng_bench_upload_row_sql()` / `ducknng_bench_upload_expected_checksum()` in `rpc_upload_compare_support.R`.
- Each writer targets its own table (`dst_w<N>`) so single-table insert contention does not swamp the protocol comparison.
- Scale up with `DUCKNNG_UPLOAD_WRITERS`, `DUCKNNG_UPLOAD_ROWS`, `DUCKNNG_UPLOAD_REPETITIONS`, `DUCKNNG_UPLOAD_TRANSPORT` environment variables.
