---
title: "ducknng bulk transport benchmark"
output:
  github_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "")
source(normalizePath("rpc_bulk_compare_support.R"))
knitr::opts_knit$set(root.dir = normalizePath(".."))

repetitions <- as.integer(Sys.getenv("DUCKNNG_BULK_REPETITIONS", unset = "5"))
rows <- ducknng_bench_parse_int_csv(
  Sys.getenv("DUCKNNG_BULK_ROWS", unset = ""),
  c(100000L, 1000000L, 10000000L)
)
raw_payload_bytes <- ducknng_bench_parse_int_csv(
  Sys.getenv("DUCKNNG_RAW_PAYLOAD_BYTES", unset = ""),
  c(1048576L, 4194304L, 16777216L)
)
db_path <- Sys.getenv(
  "DUCKNNG_BENCH_DB_PATH",
  unset = file.path(tempdir(), "ducknng_quack_tpch.duckdb")
)

results <- ducknng_bench_run_bulk_compare(
  repetitions = repetitions,
  rows = rows,
  raw_payload_bytes = raw_payload_bytes,
  db_path = db_path
)

metadata_long <- data.frame(
  field = names(results$metadata),
  value = vapply(results$metadata[1, ], function(x) as.character(x[[1]]), character(1)),
  stringsAsFactors = FALSE
)

rpc_results <- results$rpc_results[order(results$rpc_results$transport, results$rpc_results$system, results$rpc_results$rows), ]
raw_results <- results$raw_results[order(results$raw_results$transport, results$raw_results$payload_bytes), ]
ratio_results <- results$http_vs_quack[order(results$http_vs_quack$rows), ]
```

This report is rendered by `make rpc_bulk_compare`.

It measures three things:

1. `ducknng` bulk row transfer over its RPC/session surface on `http`, `tcp`, `ipc`, and `ws`.
2. `quack` bulk row transfer over its published client/server surface.
3. `ducknng` raw socket round-trip echo over `ipc`, `tcp`, and `ws`, using pair sockets without RPC, SQL session, or Arrow row framing in the transport measurement itself.

`quack` is only compared on its own public client path. `ducknng` is compared across multiple transports because transport selection is part of its SQL-facing contract.

## Machine and software details

```{r machine-details}
knitr::kable(metadata_long, col.names = c("Field", "Value"))
```

## RPC bulk transfer results

These runs execute `SELECT * FROM lineitem LIMIT n` against the server side and validate the returned data by aggregate checksum.

```{r rpc-results}
knitr::kable(rpc_results)
```

## HTTP ducknng vs Quack ratio

This is the apples-to-apples comparison on the common HTTP-facing path.

```{r http-ratio}
knitr::kable(ratio_results)
```

## Raw socket transport-only echo

These runs use `ducknng` pair sockets directly. The client sends a payload, the server echoes it back, and the client verifies the returned byte count. This isolates transport-visible round-trip cost without the RPC envelope or Arrow row encoding.

```{r raw-results}
knitr::kable(raw_results)
```

## Notes

- The row benchmark uses a local TPC-H `lineitem` table generated through DuckDB's `tpch` extension when needed.
- The raw socket benchmark uses payload sizes in bytes, not rows, because it is intentionally outside the row/RPC contract.
- `quack` is installed from `core_nightly` during the run if it is not already available.
