---
output: github_document
---

<!-- demo/motherduck_gateway.md is generated from demo/motherduck_gateway.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

library(jsonlite)
library(knitr)
library(nanonext)

project_root <- normalizePath("..", mustWork = TRUE)
extension_path <- normalizePath(
  file.path(project_root, "build", "release", "ducknng.duckdb_extension"),
  mustWork = FALSE
)
demo_script <- normalizePath(file.path(project_root, "demo", "motherduck_gateway.py"), mustWork = TRUE)

if (!file.exists(extension_path)) {
  stop("build/release/ducknng.duckdb_extension not found; run `make release` first.")
}

sql_quote <- function(text) gsub("'", "''", text, fixed = TRUE)

run_duckdb_sql <- function(code, autoload = TRUE) {
  sql <- code
  if (autoload) {
    sql <- paste(sprintf("LOAD '%s';", sql_quote(extension_path)), sql, sep = "\n")
  }
  out <- system2(
    "duckdb",
    "-unsigned",
    input = paste(".mode table", sql, sep = "\n"),
    stdout = TRUE,
    stderr = TRUE
  )
  status <- attr(out, "status")
  if (!is.null(status) && status != 0) {
    stop(paste(out, collapse = "\n"))
  }
  out
}

emit_sql_result <- function(display_code, run_code = display_code, autoload = TRUE) {
  cat("```sql\n", display_code, "\n```\n\n", sep = "")
  out <- run_duckdb_sql(run_code, autoload = autoload)
  cat("```text\n", paste(out, collapse = "\n"), "\n```\n\n", sep = "")
}

raw_hex <- function(x) paste(sprintf("%02X", as.integer(x)), collapse = "")

decode_arrow_output <- function(raw_body) {
  run_duckdb_sql(
    sprintf(
      paste(
        "SELECT *",
        "FROM ducknng_parse_body(",
        "  from_hex('%s'),",
        "  'application/vnd.apache.arrow.stream'",
        ");",
        sep = "\n"
      ),
      raw_hex(raw_body)
    )
  )
}

header_value <- function(headers, name) {
  if (is.null(headers) || is.null(names(headers))) {
    return(NA_character_)
  }
  idx <- which(tolower(names(headers)) == tolower(name))
  if (!length(idx)) {
    return(NA_character_)
  }
  as.character(headers[[idx[[1L]]]])
}

response_summary <- function(resp) {
  data.frame(
    status = resp$status,
    tenant = header_value(resp$headers, "X-Ducknng-Tenant"),
    subscriber = header_value(resp$headers, "X-Ducknng-Subscriber"),
    end_of_stream = header_value(resp$headers, "X-Ducknng-End-Of-Stream"),
    has_next_token = !is.na(header_value(resp$headers, "X-Ducknng-Next-Token")),
    stringsAsFactors = FALSE
  )
}

json_body <- function(resp) {
  if (is.null(resp$data) || !length(resp$data)) {
    return(list())
  }
  fromJSON(rawToChar(resp$data))
}

free_port <- function() {
  as.integer(system(
    "python3 -c \"import socket; s = socket.socket(); s.bind(('127.0.0.1', 0)); print(s.getsockname()[1]); s.close()\"",
    intern = TRUE
  ))
}

start_worker <- function(name, args) {
  ready_path <- tempfile(pattern = paste0("ducknng-", name, "-"), tmpdir = tempdir(), fileext = ".ready")
  pid_path <- tempfile(pattern = paste0("ducknng-", name, "-"), tmpdir = tempdir(), fileext = ".pid")
  log_path <- tempfile(pattern = paste0("ducknng-", name, "-"), tmpdir = tempdir(), fileext = ".log")
  launch_path <- tempfile(pattern = paste0("ducknng-", name, "-"), tmpdir = tempdir(), fileext = ".sh")
  launch_cmd <- paste(
    c(
      "#!/usr/bin/env bash",
      paste(
        "python3",
        shQuote(demo_script),
        paste(shQuote(args), collapse = " "),
        "--ready-file",
        shQuote(ready_path),
        ">",
        shQuote(log_path),
        "2>&1 &"
      ),
      paste("echo $! >", shQuote(pid_path))
    ),
    collapse = "\n"
  )
  writeLines(launch_cmd, launch_path)
  Sys.chmod(launch_path, mode = "700")
  status <- system2("bash", launch_path)
  if (status != 0) {
    stop("failed to launch worker ", name)
  }
  deadline <- Sys.time() + 10
  while (Sys.time() < deadline) {
    if (file.exists(ready_path)) {
      msg <- trimws(paste(readLines(ready_path, warn = FALSE), collapse = "\n"))
      if (identical(msg, "ready")) {
        return(list(name = name, ready_path = ready_path, pid_path = pid_path, log_path = log_path, launch_path = launch_path))
      }
      stop("worker ", name, " failed during startup:\n", msg)
    }
    Sys.sleep(0.1)
  }
  log_text <- if (file.exists(log_path)) paste(readLines(log_path, warn = FALSE), collapse = "\n") else ""
  stop("timed out waiting for worker ", name, "\n", log_text)
}

stop_worker <- function(worker) {
  if (is.null(worker)) {
    return(invisible(NULL))
  }
  if (file.exists(worker$pid_path)) {
    pid <- suppressWarnings(as.integer(readLines(worker$pid_path, warn = FALSE)))
    if (!is.na(pid)) {
      tools::pskill(pid)
    }
  }
  invisible(NULL)
}

wait_healthz <- function(base_url, timeout_s = 10) {
  deadline <- Sys.time() + timeout_s
  while (Sys.time() < deadline) {
    probe <- tryCatch(
      ncurl(
        paste0(base_url, "/healthz"),
        timeout = 500L
      ),
      error = function(e) NULL
    )
    if (!is.null(probe) && identical(probe$status, 200L) && identical(probe$data, "ok")) {
      return(invisible(TRUE))
    }
    Sys.sleep(0.1)
  }
  stop("gateway health check did not become ready")
}

gateway_post <- function(base_url, path, payload, api_token) {
  ncurl(
    paste0(base_url, path),
    convert = FALSE,
    response = TRUE,
    method = "POST",
    headers = c(
      "Content-Type" = "application/json",
      "Authorization" = paste("Bearer", api_token)
    ),
    data = charToRaw(toJSON(payload, auto_unbox = TRUE)),
    timeout = 2000L
  )
}

alice_token <- "demo-alice-token"
bob_token <- "demo-bob-token"
gateway_port <- free_port()
alice_backend_port <- free_port()
bob_backend_port <- free_port()
gateway_base_url <- sprintf("http://127.0.0.1:%d", gateway_port)

alice_worker <- start_worker(
  "alice-backend",
  c(
    "--role", "backend",
    "--extension", extension_path,
    "--backend-key", "alice",
    "--listen-port", as.character(alice_backend_port)
  )
)

bob_worker <- start_worker(
  "bob-backend",
  c(
    "--role", "backend",
    "--extension", extension_path,
    "--backend-key", "bob",
    "--listen-port", as.character(bob_backend_port)
  )
)

gateway_worker <- start_worker(
  "gateway",
  c(
    "--role", "gateway",
    "--extension", extension_path,
    "--gateway-port", as.character(gateway_port),
    "--subscriber", sprintf("alice=%d", alice_backend_port),
    "--subscriber", sprintf("bob=%d", bob_backend_port)
  )
)

wait_healthz(gateway_base_url)
```

# MotherDuck-style gateway demo

This is the dedicated rendered walkthrough for the multi-process subscriber gateway topology in `ducknng`. Unlike the main `README`, this document is allowed to boot several private DuckDB workers during render, so it can show the honest public-gateway shape rather than an in-process sketch.

Render it from the repo root with:

```sh
make motherduck_rdm
```

The live helper behind this document is still [demo/motherduck_gateway.py](motherduck_gateway.py). The rendered walkthrough keeps the topology generic instead of hardwiring `user -> backend` logic into the public API.

## Topology

One public DuckDB process owns the HTTP edge, identity resolution, subscriber lookup, and public continuation tokens. Several private DuckDB processes expose ordinary `ducknng` query-session services. The gateway chooses one subscriber per tenant and keeps the backend `session_id` plus `session_token` private.

```text
HTTP client
    |
    v
public gateway process
  http://127.0.0.1:<gateway>/_ducknng
    |
    +--> subscriber_alice  tcp://127.0.0.1:<alice>
    |
    +--> subscriber_bob    tcp://127.0.0.1:<bob>
```

The route layer stays generic because the gateway owns two ordinary SQL tables:

```sql
CREATE TABLE gateway_identities(
  api_token VARCHAR,
  caller_identity VARCHAR,
  tenant_id VARCHAR,
  principal_id VARCHAR,
  active BOOLEAN
);

CREATE TABLE gateway_subscribers(
  subscriber_id VARCHAR,
  tenant_id VARCHAR,
  backend_url VARCHAR,
  priority INTEGER,
  enabled BOOLEAN
);
```

`gateway_identities` resolves bearer tokens or caller identities into a tenant and principal. `gateway_subscribers` resolves that tenant into one private worker endpoint. Nothing in the public route contract needs to know whether the backend plane is two workers or two hundred.

## Gateway route shape

The public edge is still just the landed low-level HTTP route surface:

```sql
SELECT ducknng_register_http_route('gateway', 'POST', '/v1/query/start', ...);
SELECT ducknng_register_http_route('gateway', 'POST', '/v1/query/fetch', ...);
SELECT ducknng_register_http_route('gateway', 'POST', '/v1/query/close', ...);
```

Inside those routes, the gateway uses the raw session helpers so request-body columns and continuation-token columns stay dynamic:

```sql
ducknng_open_query_raw(...)
ducknng_fetch_query_raw(...)
ducknng_close_query_raw(...)
ducknng_frame_payload(...)
ducknng_frame_payload_text(...)
ducknng_frame_error_text(...)
```

That keeps the route layer transport-local and gateway-owned. The public client sees HTTP plus Arrow batches, while the private worker plane stays on the ordinary `ducknng` session contract.

## Alice starts a query through the public gateway

```r
start <- ncurl(
  paste0(gateway_base_url, "/v1/query/start"),
  convert = FALSE,
  response = TRUE,
  method = "POST",
  headers = c(
    "Content-Type" = "application/json",
    "Authorization" = "Bearer demo-alice-token"
  ),
  data = charToRaw(
    '{"sql":"SELECT owner, i, v FROM tenant_numbers ORDER BY i"}'
  ),
  timeout = 2000L
)
```

```{r alice-start, include = FALSE}
alice_start <- gateway_post(
  gateway_base_url,
  "/v1/query/start",
  list(
    sql = "SELECT owner, i, v FROM tenant_numbers ORDER BY i"
  ),
  alice_token
)
alice_next_token <- header_value(alice_start$headers, "X-Ducknng-Next-Token")
```

```{r alice-start-summary, echo = FALSE}
kable(response_summary(alice_start))
```

The response body is an Arrow IPC batch. The gateway already chose the private subscriber and exposed that choice only through headers. The effective batch split is still backend-owned, so the public route contract stays stable even when the worker chooses a larger first chunk.

```{r alice-start-body, echo = FALSE, results = "asis"}
emit_sql_result(
  paste(
    "SELECT *",
    "FROM ducknng_parse_body(<first Arrow batch>, 'application/vnd.apache.arrow.stream')",
    "LIMIT 4;",
    sep = "\n"
  ),
  sprintf(
    paste(
      "SELECT *",
      "FROM ducknng_parse_body(",
      "  from_hex('%s'),",
      "  'application/vnd.apache.arrow.stream'",
      ")",
      "LIMIT 4;",
      sep = "\n"
    ),
    raw_hex(alice_start$data)
  )
)
```

## Fetch continues on the same private subscriber

```r
fetch <- ncurl(
  paste0(gateway_base_url, "/v1/query/fetch"),
  convert = FALSE,
  response = TRUE,
  method = "POST",
  headers = c(
    "Content-Type" = "application/json",
    "Authorization" = "Bearer demo-alice-token"
  ),
  data = charToRaw(sprintf('{"token":"%s"}', alice_next_token)),
  timeout = 2000L
)
```

```{r alice-fetch, include = FALSE}
alice_fetch <- gateway_post(
  gateway_base_url,
  "/v1/query/fetch",
  list(token = alice_next_token),
  alice_token
)
alice_next_token <- header_value(alice_fetch$headers, "X-Ducknng-Next-Token")
```

```{r alice-fetch-summary, echo = FALSE}
kable(response_summary(alice_fetch))
```

```{r alice-fetch-body, echo = FALSE, results = "asis"}
emit_sql_result(
  paste(
    "SELECT *",
    "FROM ducknng_parse_body(<next Arrow batch>, 'application/vnd.apache.arrow.stream')",
    "LIMIT 4;",
    sep = "\n"
  ),
  sprintf(
    paste(
      "SELECT *",
      "FROM ducknng_parse_body(",
      "  from_hex('%s'),",
      "  'application/vnd.apache.arrow.stream'",
      ")",
      "LIMIT 4;",
      sep = "\n"
    ),
    raw_hex(alice_fetch$data)
  )
)
```

The public continuation token is gateway-owned. It carries tenant affinity and private backend session state, but it does not turn the public API into a transport-specific copy of `fetch`.

## A second identity routes to a different subscriber

```r
bob_start <- ncurl(
  paste0(gateway_base_url, "/v1/query/start"),
  convert = FALSE,
  response = TRUE,
  method = "POST",
  headers = c(
    "Content-Type" = "application/json",
    "Authorization" = "Bearer demo-bob-token"
  ),
  data = charToRaw(
    '{"sql":"SELECT owner, i, v FROM tenant_numbers ORDER BY i"}'
  ),
  timeout = 2000L
)
```

```{r bob-start, include = FALSE}
bob_start <- gateway_post(
  gateway_base_url,
  "/v1/query/start",
  list(
    sql = "SELECT owner, i, v FROM tenant_numbers ORDER BY i"
  ),
  bob_token
)
bob_next_token <- header_value(bob_start$headers, "X-Ducknng-Next-Token")
```

```{r bob-start-summary, echo = FALSE}
kable(response_summary(bob_start))
```

```{r bob-start-body, echo = FALSE, results = "asis"}
emit_sql_result(
  paste(
    "SELECT *",
    "FROM ducknng_parse_body(<bob Arrow batch>, 'application/vnd.apache.arrow.stream')",
    "LIMIT 4;",
    sep = "\n"
  ),
  sprintf(
    paste(
      "SELECT *",
      "FROM ducknng_parse_body(",
      "  from_hex('%s'),",
      "  'application/vnd.apache.arrow.stream'",
      ")",
      "LIMIT 4;",
      sep = "\n"
    ),
    raw_hex(bob_start$data)
  )
)
```

```{r bob-close-hidden, include = FALSE}
if (!is.na(bob_next_token)) {
  bob_close_hidden <- gateway_post(
    gateway_base_url,
    "/v1/query/close",
    list(token = bob_next_token),
    bob_token
  )
}
```

The public start route did not change. Only the authenticated identity changed, and the gateway routed the request to a different private subscriber.

## Cross-tenant close is rejected

```r
wrong_close <- ncurl(
  paste0(gateway_base_url, "/v1/query/close"),
  convert = FALSE,
  response = TRUE,
  method = "POST",
  headers = c(
    "Content-Type" = "application/json",
    "Authorization" = "Bearer demo-bob-token"
  ),
  data = charToRaw(sprintf('{"token":"%s"}', alice_next_token)),
  timeout = 2000L
)
```

```{r wrong-close, include = FALSE}
wrong_close <- gateway_post(
  gateway_base_url,
  "/v1/query/close",
  list(token = alice_next_token),
  bob_token
)
```

```{r wrong-close-summary, echo = FALSE}
kable(data.frame(
  status = wrong_close$status,
  body = rawToChar(wrong_close$data),
  stringsAsFactors = FALSE
))
```

That `403` is the important ownership boundary: a gateway token is not a cross-tenant escape hatch.

## The owning tenant can close the session

```r
close <- ncurl(
  paste0(gateway_base_url, "/v1/query/close"),
  convert = FALSE,
  response = TRUE,
  method = "POST",
  headers = c(
    "Content-Type" = "application/json",
    "Authorization" = "Bearer demo-alice-token"
  ),
  data = charToRaw(sprintf('{"token":"%s"}', alice_next_token)),
  timeout = 2000L
)
```

```{r right-close, include = FALSE}
right_close <- gateway_post(
  gateway_base_url,
  "/v1/query/close",
  list(token = alice_next_token),
  alice_token
)
alice_next_token <- NA_character_
```

```{r right-close-summary, echo = FALSE}
kable(data.frame(
  status = right_close$status,
  body = rawToChar(right_close$data),
  stringsAsFactors = FALSE
))
```

## What this shape buys you

This topology is generic in the right place:

- the public route surface stays fixed
- auth resolution is table-driven
- subscriber discovery is table-driven
- backend session affinity stays private
- Arrow remains the row contract

The current execution model still matters. Route handlers run on `shared_serialized_connection`, so the public gateway should stay in a separate DuckDB process from the private workers. That is exactly why this rendered document boots several processes during render instead of pretending the pattern is honest inside one runtime.

```{r cleanup, include = FALSE}
if (!is.na(alice_next_token)) {
  try(
    gateway_post(
      gateway_base_url,
      "/v1/query/close",
      list(token = alice_next_token),
      alice_token
    ),
    silent = TRUE
  )
}
stop_worker(gateway_worker)
stop_worker(bob_worker)
stop_worker(alice_worker)
```
